---
title: "Dealing with binned outcome data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Dealing with binned outcome data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
devtools::load_all()
```

In its most basic form, the version of the EM algorithm implemented in `progUkheEM` requires the following:

- a continuous outcome from *before* treatment
- a continuous outcome from *after* treatment
- an instrument affecting treatment probability but not outcomes
- a indicator for treatment.

The identification proof, however, does not need the outcomes to be continuous. This is a computing restriction. As outcomes are often wages, and wage data are often censored to some degree in surveys, here I discuss how to format binned outcome data to work with `progUkheEm`. Note that this can be tiresome if the bins/intervals are not uniform. 

We will use data from the BCS1970 survey, that contains income data in bins. Let's first load the data.

```{r load data}
data("binnedDataExample")
```

Then take a look at the first few rows, using `head`:
```{r call head}
head(binnedDataExample)
```

We can see the first column, `id`, contains unique identifiers and the second column (handily named `binnedData`) contains the binned income data. To use `progUkheEm` with binned data, we need to add two new variables for each observation: 

- a variable named `left` with the lower bound of the interval in which the true value lies
- a variable named `right` with the upper bound of the bin.

There are a number of ways to implement this. Perhaps the most straightforward is to implement this manually for each bin, which I demonstrate first. I will also show a way to automate this process for this data.

## Manually defining lower and upper bounds for each bin

First we define the variable containing the lower bound, named `left`. We use the `data.table` function `fcase`, and manually assign the value of the lower bound of the interval to `left` for each observation. The variable `binnedData` actually contain two intervals, corresponding to weekly and annual wages. We choose to use the weekly values.

```{r define lower bound}
binnedDataExample[, left := fcase(
  binnedData == "<50    pw/Under 2600 pa", 1,
  binnedData == "50-99  pw/2600-5199  pa", 50,
  binnedData == "100-149pw/5200-7799  pa", 100,
  binnedData == "150-199pw/7800-10399 pa", 150,
  binnedData == "200-249pw/10400-12999pa", 200,
  binnedData == "250-299pw/13000-15599pa", 250,
  binnedData == "300-349pw/15600-18199pa", 300,
  binnedData == "350-399pw/18200-20799pa", 350,
  binnedData == "400-449pw/20800-23399pa", 400,
  binnedData == "450-499pw/23400-25999pa", 450,
  binnedData == "500&over /26000&over", 500
)]

head(binnedDataExample)
```

Next we define the upper bound, `right`, in exactly the same way. 

```{r define upper bound}
binnedDataExample[, right := fcase(
  binnedData == "<50    pw/Under 2600 pa", 50,
  binnedData == "50-99  pw/2600-5199  pa", 99,
  binnedData == "100-149pw/5200-7799  pa", 149,
  binnedData == "150-199pw/7800-10399 pa", 199,
  binnedData == "200-249pw/10400-12999pa", 249,
  binnedData == "250-299pw/13000-15599pa", 299,
  binnedData == "300-349pw/15600-18199pa", 349,
  binnedData == "350-399pw/18200-20799pa", 399,
  binnedData == "400-449pw/20800-23399pa", 449,
  binnedData == "450-499pw/23400-25999pa", 499,
  binnedData == "500&over /26000&over", 1500
)]

head(binnedDataExample)
```

## Automating assignment of lower and upper

Here I present a method of automating the definitions of `left` and `right` from the binned data. The exact code will depend on the form the intervals/bins are recorded in.

First we need to isolate the key information from the variable identifying the bin/interval. Here it is the part of the string preceding the "/" that defines the interval for weekly wages. The variable containing the intervals are recorded as factors. 

```{r extract key info on interval}
# first rename manual variables
setnames(binnedDataExample, 
         old = c("left", "right"), 
         new = paste0(c("left", "right"), "_man"))

# check type of binnedData
binnedDataExample[, is.factor(binnedData)]

# extract levels
binnedData_levels <- binnedDataExample[, levels(binnedData)]

# select relevant info from string
keyInfo <- unlist(stringr::str_split_fixed(binnedData_levels, "/", n = 2))[, 1]
```

We now have the relevant information stored in `keyInfo`. We now want to extract the upper and lower bound to separate variables. We will use `str_extract_all` from the `stringr` package and regular expressions.

```{r }
lower <- c(
  1, 
  as.numeric(stringr::str_extract_all(keyInfo, 
                                      "[0-9]{2,}", 
                                      simplify = TRUE)[2:13, 1])
)
upper <- c(
  50, 
  as.numeric(stringr::str_extract_all(keyInfo, 
                                      "[0-9]{2,}", 
                                      simplify = TRUE)[2:10, 2]),
  1500
)
```

Assign `left` and `right`.

```{r assign left and right}
binnedDataExample[, ':='(
  left = lower[as.numeric(binnedData)],
  right = upper[as.numeric(binnedData)]
)]

head(binnedDataExample)
```

We can check that the manually assigned and automated variables are the same:
```{r check man and auto agree}
binnedDataExample[, all.equal(left_man, left)]

binnedDataExample[, all.equal(right_man, right)]
```
