---
title: "desc_analysis_bcs"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{desc_analysis_bcs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
# library(ukheEm)
devtools::load_all()

library(ggplot2)
library(magrittr)
```

## Summary statistics

```{r observed characteristics}

dtBcsWorkerFirm <- bcs70$bcs1996x[, .(
  bcsid, sex, degree,
  yrJobStart = b960270,
  inCouple = b960319,
  married = b960322,
  livingSitn = b960321,
  nbKids = b960333,
  age = b960338,
  health = b960432,
  soc, seg91, rgsc91, isco, goldth90, hgscale, # occupation / socio-econ classes
  noVocQual = no_qual,
  ol_ac, ol_de, gcse_ac, gcse_de, # O-level / GCSE grades
  alevel, slevel, # A-level / S-level passes,
  hed, pgce, opgc, postgrad, found, dip_un, other_ac, # academic qualifications
  nvq_1, nvq_2, nvq_3, nvq_4, nvq_5, nvq_6, # nvqs at levels
  ageLeftSchl = b960129, ageLeftFtEd = b960132,
  trainCourse = b960134, yts = b960137, employerTrain = b960143, # traininig since end fte
  lengthUnemp = b960258, 
  usualHoursWk = b960277, 
  firmSize = b960272
)]

dtBcs <- merge(
  dtBcs4Em, 
  dtBcsWorkerFirm, 
  by = "bcsid"
) %>% 
  merge(bcs70$bcs1986x[, .(bcsid, ethnicGroup = c6.14)])

dtBcs <- dtBcs[complete.cases(dtBcs[, .(left, logWkPay, cogScore, c5e7, degree.x)])]
```

```{r summary stats BCS}
dtBcsSumStats <- dtBcs[, .(
  "Par. Inc. @16 (mean)" = mean(exp(left+right)/52),
  "Par. Inc. (std dev.)" = sd(exp(left+right)/52),
  "Wage (mean)" = mean(exp(logWkPay)),
  "Wage (std dev.)" = sd(exp(logWkPay)),
  "Degree" = mean(degree.x),
  "Male" = mean(sex %like% "Male"),
  "In relationship" = mean(inCouple %like% "Yes"),
  "Married" = mean(married %in% c("Married - 1st", "Remarried - 2nd+")),
  "Divorced" = mean(married %in% c("Separated", "Divorced")),
  "Live w/ partner" = mean(livingSitn %in% c("Living with spouse", 
                                             "Living as a couple")),
  "Has children" = mean(nbKids %in% paste0(1:5)),
  "Num. of children" = mean(as.numeric(paste0(nbKids)), na.rm = TRUE), 
  "In good health" = mean(health %in% c("Excellent", "Good")),
  "Occupation data" = mean(!is.na(hgscale)),
  "H-G score" = mean(hgscale, na.rm = TRUE),
  "Age left schl" = mean(as.numeric(paste0(ageLeftSchl)), na.rm = TRUE), 
  "Age left ft ed." = mean(as.numeric(paste0(ageLeftFtEd)), na.rm = TRUE),
  "Ever unemp." = mean(lengthUnemp %in% c(
    "3 months or less", "4-6 months", "7-11 months", "1-2 years", 
    "More than 2 years"
  )),
  "Hours per week" = mean(as.numeric(paste0(usualHoursWk)), na.rm = TRUE),
  "10 or less" = mean(firmSize %like% "10 or less"), 
  "11--25" = mean(firmSize %like% "11 - 25"),
  "26--99" = mean(firmSize %like% "26 - 99"),
  "100--499" = mean(firmSize %like% "100 - 499"),
  "500 or more" = mean(firmSize %like% "500 or more"),
  "Self-esteem" = mean(selfEsteemScore, na.rm = TRUE),
  "Loc" = mean(locScore, na.rm = TRUE),
  "GHQ" = mean(ghqScore, na.rm = TRUE),
  "Cognitive" = mean(cogScore),
  "Noncognitive score" = mean(noncogScore, na.rm = TRUE),
  "Reading score" = mean(readScore, na.rm = TRUE),
  "Maths score" = mean(mathScore, na.rm = TRUE),
  "Leave home" = mean(c5e7 %in% c("Matters very much", "Matters somewhat")),
  "Leave home (matters alot)" = mean(c5e7 %in% c("Matters very much")),
  "Leave home (matters somewhat)" = mean(c5e7 %in% c("Matters somewhat")),
  "Leave home (doesn't matter)" = mean(c5e7 %in% c("Doesn't matter")),
  .N
)] %>%
  # .[, names(.) := lapply(.SD, signif2), 
  #            .SDcols = names(.)] %>%
  t() %>%
  as.data.table(keep.rownames = TRUE) 

xtable::xtable(dtBcsSumStats, digits = 2)

dtBcs[degree.x == FALSE, .(prop = .N / 1290), by = rgsc91][order(rgsc91)]
dtBcs[degree.x == TRUE, .(prop = .N / 522), by = rgsc91][order(rgsc91)]

dtBcs[degree.x == FALSE, .(prop = .N / 1290), by = ethnicGroup]
dtBcs[degree.x == TRUE, .(prop = .N / 522), by = ethnicGroup]


dtBcs[, .(prop = .N / 1812), by = ethnicGroup]

dtBcs[, soc9 := substr(soc, 1, 1)][, soc4 := fcase(
  soc9 %in% c("1", "2"), "HigherMan",
  soc9 %in% c("3", "5"), "IntermOcc",
  soc9 %in% c("4", "6", "7", "8"), "LowerSuperv",
  soc9 == "9", "Semi-routine"
)]

dtBcs[, .N / NN, by = soc4]

dtBcs[degree.x == F, .N / 1290, by = soc4]
dtBcs[degree.x == T, .N / 522, by = soc4]
```

## OLS wage model

```{r ols wage}

dtBcs[, parIncContn := left + right / 2]

wageModelBcs <- lm(
  logWkPay ~ parIncContn + combnScore + degree.x + sex + ethnicGroup +
    selfEsteemScore + locScore + ghqScore, 
  data = dtBcs
)

summary(wageModelBcs)


```

```{r lasso to select best vars}

oldNpVars <- c(
  # about work
  "c5a1", "c5a2", "c5a3", "c5a4", "c5a5", "c5a6", "c5a7", "c5a8", "c5a9",
  # job related
  "c5d1", "c5d2", "c5d3", "c5d4", "c5d5", "c5d6", "c5d7", "c5d8",
  "c5d9", "c5d10", "c5d11", "c5d12", "c5d13", "c5d14", "c5d15", "c5d16",
  # adult life
  "c5e1", "c5e2", "c5e3", "c5e4", "c5e5", "c5e6", "c5e7", "c5e8", "c5e9",
  "c5e10", "c5e11", "c5e12", "c5e13", "c5e14", "c5e15"
)

newNpVars <- c(
  # about work
  "qualHelpJob", "whoKnowHelpJob", "deterHelpJob", "luckGetJob",
  "educOnlyDelayEmp", "bestLvSchlASAP", "planCareerUseless", "takeAnyJob",
  "expGrQuals",
  # job related
  "helpOth", "hiPay", "goodBoss", "wrkOutside", "selfEmp", "jobVaries",
  "jobEasy", "jobAmbi", "jobMath", "tradeProf", "jobQuiet", "jobLT", "jobChall",
  "jobTravel", "jobBuild", "jobRegHrs",
  # adult life
  "moreFun", "ftJob", "selfResp", "notBossed", "canVote", "noDoss", "moveOut",
  "getMarr", "nightClub", "localComm", "xFilms", "drinkAlc", "politics", "kids",
  "doWhatIWant"
  # personality ones to add
)

setnames(dtBcs, oldNpVars, newNpVars)

lassoFormulaBcs <- as.formula(
  paste0("logWkPay ~ ", paste0(
    c("parIncContn", "combnScore", "degree.y", "sex", "ethnicGroup", 
      "selfEsteemScore", "locScore", "ghqScore", newNpVars), collapse = "+"
  ))
)

wageModelBcs2 <- lm(
  lassoFormulaBcs,
  data = dtBcs
)

coefplot::coefplot(wageModelBcs2, sort = "magnitude")

lassoDt <- model.frame(
  lassoFormulaBcs, data = dtBcs
)
lassoX <- model.matrix(
  lassoFormulaBcs, data = dtBcs
)

lassoY <- lassoDt[, "logWkPay"]

library(glmnet)

lassoWageBcs <- cv.glmnet(
  x = lassoX, y = lassoY
)

plot(lassoWageBcs)

# coefficients for "best" lambda
coef(lassoWageBcs, s = "lambda.min")

# coefficients for "one-standard-error" lambda
coef(lassoWageBcs, s = "lambda.1se")

# coefficients for other lambda
coef(lassoWageBcs, s = exp(-3.4))
```

## Test scores ($y_1$)

```{r test scores}
dtScores <- dtBcs4Em[, .(bcsid, mathScore, readScore)]

corrMathRead <- dtScores[, cor(mathScore, readScore, 
                               use = "na.or.complete")]

# combine maths and reading score
dtScores[, combnScore := fcase(
  !is.na(mathScore) & !is.na(readScore), mathScore + readScore / 2,
  !is.na(readScore) & is.na(mathScore), readScore,
  !is.na(mathScore) & is.na(readScore), mathScore
)]

# combined score no missing
dtScores[, combScore_true := mathScore + readScore / 2]

dtScores <- melt(dtScores, id.vars = "bcsid", 
                 value.name = "score")

ggplot(data = dtScores, 
       mapping = aes(x = score, colour = variable)) +
  geom_density() +
  scale_color_discrete(name = NULL, 
                       limits = c("readScore", "mathScore", 
                                  "combnScore", "combnScore_true"),
                       labels = c("Reading", "Maths", "Combined", "Combined (no missing)")) +
  annotate(geom = "label", x = 7, y = .028, label.size = NA,
           label = paste0("cor(R, M) = ", 
                          signif(corrMathRead, digits = 3))) +
  theme_bw() + 
  theme(legend.position = c(.1, .9))
```

## Other measures of (cognitive) ability (for comparison)

```{r other measures ability}

dtGrades <- bcs70$bcs1986x[, .(
  bcsid, 
  engLangOL = t2a1.2,
  engLitOL = t2b1.2,
  mathsOL = t2c1.2,
  sciOL = t2d1.2,
  physOL = t2e1.2,
  bioOL = t2d1.2,
  engLangCSE = t2a2.2,
  engLitCSE = t2b2.2,
  mathsCSE = t2c2.2,
  sciCSE = t2d2.2,
  physCSE = t2e2.2,
  bioCSE = t2d2.2
)]

colsOL <- names(dtGrades)[names(dtGrades) %like% "OL"]
colsCSE <- names(dtGrades)[names(dtGrades) %like% "CSE"]

dtGrades[, paste0(colsOL, "_AC") := lapply(.SD, function(x) x %in% c("Grade A", "Grade B", "Grade C")),
         .SDcols = colsOL]

dtGrades[, paste0(colsCSE, "_AC") := lapply(.SD, function(x) x %in% c("Grade 1")),
         .SDcols = colsCSE]

dtGrades[, numA_C := rowSums(.SD), .SDcols = paste0(c(colsOL, colsCSE), "_AC")]

dtGradesScores <- merge(dtScores, dtGrades[, .(bcsid, numA_C)], by = "bcsid")


ggplot(data = dtGradesScores[variable == "combnScore"], 
       mapping = aes(x = score, fill = factor(numA_C))) +
  geom_histogram() +
  scale_fill_brewer(name = "GCSEs A*-C", palette = "YlOrRd") +
  theme_bw() + 
  theme(legend.position = c(.1, .75)) +
  guides(fill = guide_legend(ncol = 2))

dtGrades[, table(numA_C)]
```

## Measures of noncognitive abilities

- Right and might [5B]
- Compared with others [5G]
- How I feel [5I] (GHQ)
- Fate and fortune [5L] (Caraloc)
- Feeling healthy [5O] (Malaise)
- Self esteem [F22] 

```{r NC measures}

locVarsN <- paste0("c5l", c(1:3, 5, 7, 8, 10, 12:15, 17, 19, 20, 23, 25, 26))
locVarsY <- paste0("c5l", c(9, 22))
ghqVarsP <- paste0("c5i", 1:6)
ghqVarsN <- paste0("c5i", 7:12)

cols2keep <- c("bcsid", "f22score", locVarsN, locVarsY, ghqVarsN, ghqVarsP)

dtNonCog <- bcs70$bcs1986x[, ..cols2keep] %>%
  .[, selfEsteemScore := as.numeric(f22score)] %>%
  .[, locScoreN := rowSums(.SD == "No"), .SDcols = locVarsN] %>% 
  .[, locScoreY := rowSums(.SD == "Yes"), .SDcols = locVarsN] %>% 
  .[, locScore := locScoreN + locScoreY] %>%
  .[, ghqScoreN := rowSums(sapply(.SD, as.numeric) - 1), .SDcols = ghqVarsN] %>%
  .[, ghqScoreP := rowSums((sapply(.SD, as.numeric) * -1) + 4), .SDcols = ghqVarsP] %>%
  .[, ghqScore := ghqScoreP + ghqScoreN]

ggplot(data = dtNonCog) +
  geom_histogram(aes(x = locScore, colour = "LoC", fill = "LoC"), bins = 20) +
  geom_histogram(aes(x = selfEsteemScore, colour = "Self-esteem", fill = "Self-esteem"), bins = 20) +
  scale_fill_manual(name = "NC measure", limits = c("LoC", "Self-esteem"), values = c("red", NA)) +
  scale_colour_manual(name = "NC measure", limits = c("LoC", "Self-esteem"), values = c(NA, "blue")) +
  scale_x_continuous(name = "Score") +
  theme_classic() +
  theme(legend.position = c(.15, .85), text = element_text(size = 15))

ggplot(data = dtNonCog[selfEsteemScore > 7 & locScore > 7],
       mapping = aes(x = locScore, y = selfEsteemScore)) +
  geom_density_2d_filled() + 
  theme_classic()

x <- dtNonCog[!is.na(selfEsteemScore) & !is.na(locScore) & !is.na(ghqScore), 
              locScore]
y <- dtNonCog[!is.na(selfEsteemScore) & !is.na(locScore) & !is.na(ghqScore), 
              selfEsteemScore]
z <- dtNonCog[!is.na(selfEsteemScore) & !is.na(locScore) & !is.na(ghqScore), 
              ghqScore]

den3d <- MASS::kde2d(x, y)

persp(den3d)

library(plotly)
plot_ly(x = den3d$x, y = den3d$y, z = den3d$z) %>% 
  add_surface() %>%
  layout(
    title = "Density of noncognitive scores",
    scene = list(
      xaxis = list(title = "Locus of control"),
      yaxis = list(title = "Self-esteem"),
      zaxis = list(title = "Density")
    )
  )

ggplot(data = dtNonCog,
       mapping = aes(x = locScore, y = selfEsteemScore)) +
  geom_jitter() +
  theme_classic()

ggplot(data = dtNonCog,
       mapping = aes(x = locScore, y = ghqScore)) +
  geom_jitter() +
  theme_classic()

ggplot(data = dtNonCog,
       mapping = aes(x = ghqScore, y = selfEsteemScore)) +
  geom_jitter() +
  theme_classic()
```

### PCA on NC scores

```{r}

library(factoextra)

resPca <- prcomp(
  dtNonCog[!is.na(selfEsteemScore) & !is.na(locScore) & !is.na(ghqScore), 
           .(locScore, selfEsteemScore, ghqScore)],
  scale = TRUE
)

fviz_eig(resPca)

fviz_pca_var(
  resPca,
  col.var = "contrib", # Color by contributions to the PC
  gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
  repel = TRUE     # Avoid text overlapping
)
```

## (Weekly) pay at twenty-five ($y_2$)

```{r pay at 25}

ggplot(data = dtBcs4Em, 
       mapping = aes(x = wkPay, colour = degree)) +
  geom_density() +
  theme_classic()

ggplot(data = dtBcs4Em, 
       mapping = aes(x = logWkPay, colour = degree)) +
  geom_density() +
  theme_classic()


dtBcs4Em[, median(wkPay), by = degree]

```


## Possible instruments ($z$)

### What about work? [5a]

```{r choosing an instrument waw}

dtBcs4Em[, names(dtBcs4Em)[startsWith(names(dtBcs4Em), "c5a")] := lapply(
  .SD, relevel, ref = "Disagree"), 
  .SDcols = names(dtBcs4Em)[startsWith(names(dtBcs4Em), "c5a")]]

degreeLogitZFormula_waw <- as.formula(
  paste0("degree ~ ", 
         paste0(names(dtBcs4Em)[startsWith(names(dtBcs4Em), "c5a")], 
                collapse = " + "))
)

degreeLogitZ_waw <- glm(
  formula = degreeLogitZFormula_waw,
  data = dtBcs4Em,
  family = binomial()
)

dtDegreeLogitZ_waw <- as.data.table(
  summary(degreeLogitZ_waw)$coefficients, 
  keep.rownames = TRUE
)

dtDegreeLogitZ_waw[, c("varNames", "answer") := tstrsplit(rn, "A")]
dtDegreeLogitZ_waw[!is.na(answer), answer := paste0("A", answer)]

dtZ_waw <- bcs70labels$bcs1986x[varNames %like% c("c5a"), ]

dtDegreeLogitZ_waw <- merge(
  dtDegreeLogitZ_waw, dtZ_waw, by = "varNames"
)

dtDegreeLogitZ_waw[, ci05 := Estimate - 1.96 * `Std. Error`]
dtDegreeLogitZ_waw[, ci95 := Estimate + 1.96 * `Std. Error`]

```

```{r plot degreelogit waw coefs}

ggplot(
  data = dtDegreeLogitZ_waw,
  mapping = aes(x = varLabels, y = Estimate, 
                ymin = ci05, ymax = ci95,
                colour = answer)
) + 
  geom_hline(yintercept = 0, colour = "grey") +
  geom_pointrange() +
  scale_color_discrete(name = "Response  [ref. cat. Disagree]") +
  theme_classic() +
  coord_flip() + theme(legend.position = "bottom") +
  labs(subtitle = "What about work?")

```

### What's in a job? [5d]

```{r choosing an instrument jobs}

dtBcs4Em[, names(dtBcs4Em)[startsWith(names(dtBcs4Em), "c5d")] := lapply(
  .SD, relevel, ref = "Doesn't matter"), 
  .SDcols = names(dtBcs4Em)[startsWith(names(dtBcs4Em), "c5d")]]

degreeLogitZFormula_job <- as.formula(
  paste0("degree ~ ", 
         paste0(names(dtBcs4Em)[startsWith(names(dtBcs4Em), "c5d")], 
                collapse = " + "))
)

degreeLogitZ_job <- glm(
  formula = degreeLogitZFormula_job,
  data = dtBcs4Em,
  family = binomial()
)

dtDegreeLogitZ_job <- as.data.table(
  summary(degreeLogitZ_job)$coefficients, 
  keep.rownames = TRUE
)

dtDegreeLogitZ_job[, c("varNames", "answer") := tstrsplit(rn, "M")]
dtDegreeLogitZ_job[!is.na(answer), answer := paste0("M", answer)]

dtZ_job <- bcs70labels$bcs1986x[varNames %like% c("c5d"), ]

dtDegreeLogitZ_job <- merge(
  dtDegreeLogitZ_job, dtZ_job, by = "varNames"
)

dtDegreeLogitZ_job[, ci05 := Estimate - 1.96 * `Std. Error`]
dtDegreeLogitZ_job[, ci95 := Estimate + 1.96 * `Std. Error`]

```

```{r plot degreelogit coefs jobs}

ggplot(
  data = dtDegreeLogitZ_job,
  mapping = aes(x = varLabels, y = Estimate, 
                ymin = ci05, ymax = ci95,
                colour = answer)
) + 
  geom_hline(yintercept = 0, colour = "grey") +
  geom_pointrange() +
  scale_x_discrete(
    name = NULL,
    label = as_labeller(function(x) stringr::str_remove(x, "Job: "))
  ) +
  scale_color_discrete(name = "Response [ref. cat. Doesn't matter]") +
  theme_classic() +
  coord_flip() + theme(legend.position = "bottom") +
  labs(subtitle = "What's in a job?")

```


### Looking ahead [5e] (about becoming an adult)

```{r choosing an instrument}

dtBcs4Em[, names(dtBcs4Em)[startsWith(names(dtBcs4Em), "c5e")] := lapply(
  .SD, relevel, ref = "Doesn't matter"), 
  .SDcols = names(dtBcs4Em)[startsWith(names(dtBcs4Em), "c5e")]]

degreeLogitZFormula_adult <- as.formula(
  paste0("degree ~ ", 
         paste0(names(dtBcs4Em)[startsWith(names(dtBcs4Em), "c5e")], 
                collapse = " + "))
)

degreeLogitZ_adult <- glm(
  formula = degreeLogitZFormula_adult,
  data = dtBcs4Em,
  family = binomial()
)

dtDegreeLogitZ_adult <- as.data.table(
  summary(degreeLogitZ_adult)$coefficients, 
  keep.rownames = TRUE
)

dtDegreeLogitZ_adult[, c("varNames", "answer") := tstrsplit(rn, "M")]
dtDegreeLogitZ_adult[!is.na(answer), answer := paste0("M", answer)]

dtZ_adult <- bcs70labels$bcs1986x[varNames %like% c("c5e"), ]

dtDegreeLogitZ_adult <- merge(
  dtDegreeLogitZ_adult, dtZ_adult, by = "varNames"
)

dtDegreeLogitZ_adult[, ci05 := Estimate - 1.96 * `Std. Error`]
dtDegreeLogitZ_adult[, ci95 := Estimate + 1.96 * `Std. Error`]

```

```{r plot degreelogit coefs}

ggplot(
  data = dtDegreeLogitZ_adult,
  mapping = aes(x = varLabels, y = Estimate, 
                ymin = ci05, ymax = ci95,
                colour = answer)
) + 
  geom_hline(yintercept = 0, colour = "grey") +
  geom_pointrange() +
  scale_x_discrete(
    name = NULL,
    label = as_labeller(function(x) stringr::str_remove(x, "Adult:? ?l?i?f?e?:? ?"))
  ) +
  scale_color_discrete(name = "Response [ref. cat. Doesn't matter]") +
  theme_classic() +
  coord_flip() + theme(legend.position = "bottom") +
  labs(subtitle = "Looking ahead [to adult life]")

```

## MCA on questions above

Multiple correspondance analysis (MCA) is a technique similar to principal components analysis (PCA) that is applicable to categorical variables. See vignette `FactoMineR_MCA` for more details. 

```{r mca}

library(FactoMineR)

cols4mca <- c(names(dtBcs4Em)[names(dtBcs4Em) %like% "c5"], "degree", "wkPay")

dt4Mca <- dtBcs4Em[, ..cols4mca][complete.cases(dtBcs4Em[, ..cols4mca]), ]

resMca <- MCA(dt4Mca, 
              quali.sup = grep(names(dt4Mca), pattern = "degree"),
              quanti.sup = grep(names(dt4Mca), pattern = "wkPay"))

```

```{r plot of individuals}

plot(
  resMca, choix = "ind",
  invisible = c("var"),
  cex = .5, label = "none",
  title = "Scatter plot of individuals on first two dimensions",
  habillage = "degree"
)

```

```{r plot active vars}

plot(resMca, # choix = "var", 
     invisible = c("ind", "quali.sup"),
     col.var = c(
       rep("red", 27),    # c5axx "What about work?"
       rep("black", 48),  # c5dxx "What's in a job?"
       rep("blue", 45)      # c5exx "Looking ahead"
     ),
     title = "Graph of the active categories",
     xlim = c(-2, 2), label = "none")

dimsResMca <- dimdesc(resMca)

dtDim1 <- as.data.table(dimsResMca$`Dim 1`$category, keep.rownames = TRUE)

```

- all variables

```{r all vars mca}

plot(resMca, choix = "var", 
     title = "Scatter plot of all variables",
     label = "none")

```
