---
title: "descriptiveStats"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{descriptiveStats}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
# library(ukheEm)
load_all()
library(ggplot2)

# load data
data("lsype1YP", "lsype1FB", "lsype1YPlabels", "lsype1FBlabels")

# create dtLsype
lsype1YP <- lapply(lsype1YP, as.data.table)

benefitCols <- names(lsype1YP[[4]])[names(lsype1YP[[4]]) %like% "Benefits"]
costCols    <- names(lsype1YP[[4]])[names(lsype1YP[[4]]) %like% "Costs"]

WhyHeCols <- names(lsype1YP[[4]])[names(lsype1YP[[4]]) %like% "WHYHEYP"]
HeConCols <- names(lsype1YP[[4]])[names(lsype1YP[[4]]) %like% "HeCon"]

# new variable indicating if asked b/c questions
lsype1YP[[4]][, inSubSample := !is.na(W4BenefitsYP0a)]

dtLsype <- merge(
  lsype1YP[[4]][, .(NSID, female = (W4SexYP == "Female"), ethGrp = W4ethgrpYP,
                    mainAct17 = W4MainActYP, inSubSample, 
                    plan2applyHe = fcase(
                      W4Heposs9YP %in% c("Very likely", "Fairly likely"), T,
                      W4Heposs9YP %in% c("Not very likely", 
                                         "Not at all likely"), F
                    ),
                    .SD), 
                .SDcols = c(benefitCols, costCols, WhyHeCols, HeConCols)], 
  lsype1YP[[8]][, .(NSID, W8USUAL, mainAct25 = W8DACTIVITY, 
                    employed25 = fcase(W8DWRK == "Currently employed", T,
                                       W8DWRK == "Not currently employed", F), 
                    degree = fcase(W8DDEGP == "First or higher degree", T,
                                   W8DDEGP == "No degree", F),
                    russellGrp = fcase(
                      W8DRUSSELL == "Awarded by Russell Group university", T,
                      W8DRUSSELL == "Awarded by other HE institution", F),
                    wage25 = W8GROW)],
  on = "NSID"
)

setnames(dtLsype, 
         old = paste0(".SD.", c(benefitCols, costCols, WhyHeCols, HeConCols)),
         new = c(benefitCols, costCols, WhyHeCols, HeConCols))

```

This vignette produces descriptive statistics and plots for the datasets I use in the UKHE project(s).

# Next Steps (LSYPE1)

The main data source for this project is the longitudinal cohort study Next Steps, previously the Longitudinal Study of Young People in England 1 (LSYPE1).

## Descriptive statistics: full vs subsample

I select my subsample based on some key questions that were only asked to some of the students. 
These questions relate to the advantages and disadvantages of attending university, and were on asked to students who had at least 5 GCSEs at grade A*-C or passes at intermediate GNVQ. 
This does not seem to be an official "entry standard" to university, and it is not entirely clear to me why this selection was made for this question. 

```{r desc stats lsype1}

dtLsype[, .(
  .N, 
  Female = mean(female, na.rm = T), 
  University = mean(degree, na.rm = T),
  RussellGroup = mean(russellGrp, na.rm = T),
  Employed25 = mean(employed25, na.rm = T),
  Wage25 = median(wage25, na.rm = T)
), by = inSubSample]

dtLsype[, .(
  .N, 
  Female = mean(female, na.rm = T), 
  University = mean(degree, na.rm = T),
  RussellGroup = mean(russellGrp, na.rm = T),
  Employed25 = mean(employed25, na.rm = T),
  Wage25 = median(wage25, na.rm = T)
)]

```
### Wage distributions

```{r earning dist lsype}

wageQs <- dtLsype[, quantile(wage25, probs = c(0, .99), na.rm = T)]

ggplot(
  data = dtLsype[!is.na(degree)],
  mapping = aes(x = wage25, colour = degree, linetype = degree) 
) +
  geom_line(stat = "density") +
  scale_x_continuous(name = "Wage @25 (gross weekly, GBP)", 
                     limits = wageQs) +
  scale_color_discrete(name = "First degree") +
  scale_linetype_discrete(name = "First degree") +
  theme_bw() +
  theme(legend.position = c(.9, .9), 
        text = element_text(size = 15))

# ggsave(
#   filename = "/home/opmc/Dropbox/work/UkHeExp/tex/fig/ReducedForm/wageDistsLsype.png",
#   width = 10, height = 6
# )

```

### Other factors ("psychic costs")

```{r other factor questions lsype}

# benefit labels
benefitLabels <- c(
  "Get better job", "Well-paid job", "Better opportunities", "Need for career",
  "Show skills", "Delay get job", "Social life", "Leave home", "Learning", 
  "More qualifications", "Personal development", "More confidence",
  "More respect", "Better life (general)", "Gain life skills", "Other", 
  "Don't know", "No answer"
)

costLabels <- c(
  "Expensive", "Get into debt", "Depend on parents", "Not financially indep.",
  "Not earning / working", "Costs (general)", "No job guarantee", 
  "Not needed for job", "Less experience", "Heavy workload", "Leave home",
  "Takes long time", "Waste of time", "Tuition fees etc.", "Stress", "Other",
  "Don't know", "No answer"
)

dtBenefits <- dtLsype[inSubSample == T, 
                            .(benefit = names(.SD), 
                              count = colSums(.SD == "Mentioned")), 
                            .SDcols = benefitCols, 
                      by = c("plan2applyHe", "degree")]

dtCosts <- dtLsype[inSubSample == T, 
                            .(cost = names(.SD), 
                              count = colSums(.SD == "Mentioned")), 
                            .SDcols = costCols, 
                   by = c("plan2applyHe", "degree")]

# plot and save benefits
ggplot(
  data = dtBenefits,
  mapping = aes(x = benefit, y = count, fill = plan2applyHe)
) +
  geom_col() +
  scale_x_discrete(name = NULL, limits = benefitCols, labels = benefitLabels) +
  scale_fill_discrete(name = "Plan to:", 
                      limits = c(T, F, NA), labels = c("Apply to HE",
                                                       "Not apply to HE", 
                                                       "Unsure / other")) +
  coord_flip() +
  theme_bw() +
  theme(text = element_text(size = 15), legend.position = c(.85, .85))

ggsave(
  filename = "/home/opmc/Dropbox/work/UkHeExp/tex/fig/ReducedForm/benefitsPlot.png",
  width = 10, height = 6
)

# plot and save costs
ggplot(
  data = dtCosts,
  mapping = aes(x = cost, y = count, fill = plan2applyHe)
) +
  geom_col() +
  scale_x_discrete(name = NULL, limits = costCols, labels = costLabels) +
  scale_fill_discrete(name = "Plan to:", 
                      limits = c(T, F, NA), labels = c("Apply to HE",
                                                       "Not apply to HE", 
                                                       "Unsure / other")) +
  coord_flip() +
  theme_bw() +
  theme(text = element_text(size = 15), legend.position = c(.85, .85))

ggsave(
  filename = "/home/opmc/Dropbox/work/UkHeExp/tex/fig/ReducedForm/costsPlot.png",
  width = 10, height = 6
)
```

```{r main reasons} 

WhyHeCols <- names(lsype1YP[[4]])[names(lsype1YP[[4]]) %like% "WHYHEYP"]
HeConCols    <- names(lsype1YP[[4]])[names(lsype1YP[[4]]) %like% "HeCon"]

WhyHeLabels <- c(
  "Get better job", "Well-paid job", "Better opportunities", "Need for career",
  "Show skills", "Delay get job", "Learning", "Get a degree",
  "More qualifications", "Social life", "Leave home", "Personal development", 
  "More confidence", "More respect", "Expected by teachers", 
  "Expected by fam/friends", "Better life (general)", "Gain life skills", 
  "Other", "No answer", "Don't know", "Refusal"
)

HeConLabels <- c(
  "Expensive", "Get into debt", "Depend on parents", "Not financially indep.",
  "Not earning / working", "Costs (general)", "No job guarantee", 
  "Teacher advised get job", "Family advised get job", 
  "Not needed for job", "Application difficult", "Not got grades",
  "Break from study", "Not sure I'd like", "Not interested", "Not fit in",
  "Heavy workload", "Friends got jobs", "No particular reason",
  "Takes long time", "Waste of time", "Other",
  "Don't know", "Refusal"
)

# count number of answers
dtWhyHe <- dtLsype[inSubSample == T, 
                            .(reason = names(.SD), 
                              count = colSums(.SD == "Mentioned", na.rm = T)), 
                            .SDcols = WhyHeCols, by = degree]

dtHeCon <- dtLsype[inSubSample == T, 
                            .(reason = names(.SD), 
                              count = colSums(.SD == "Mentioned", na.rm = T)), 
                            .SDcols = HeConCols, by = degree]

# plot and save why HE
ggplot(
  data = dtWhyHe,
  mapping = aes(x = reason, y = count, fill = degree)
) +
  geom_col() +
  scale_x_discrete(name = NULL, limits = WhyHeCols, labels = WhyHeLabels) +
  scale_fill_discrete(name = NULL, limits = c(T, F, NA), 
                      labels = c("BA +", "No degree", NA)) +
  coord_flip() +
  theme_bw() +
  theme(text = element_text(size = 15), legend.position = c(.85, .85))

ggsave(
  filename = "/home/opmc/Dropbox/work/UkHeExp/tex/fig/ReducedForm/whyHePlot.png",
  width = 10, height = 6
)

# plot and save costs
ggplot(
  data = dtHeCon,
  mapping = aes(x = reason, y = count, fill = degree)
) +
  geom_col() +
  scale_x_discrete(name = NULL, limits = HeConCols, labels = HeConLabels) +
  scale_fill_discrete(name = NULL, limits = c(T, F, NA), 
                      labels = c("BA +", "No degree", NA)) +
  coord_flip() +
  theme_bw() +
  theme(text = element_text(size = 15), legend.position = c(.85, .85))

ggsave(
  filename = "/home/opmc/Dropbox/work/UkHeExp/tex/fig/ReducedForm/heConPlot.png",
  width = 10, height = 6
)

```
