---
title: "An initial analysis of the data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{initialAnalysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
# library(ukheEm)
# devtools::load_all()
```

This vignette documents the "reduced-form" analysis that the EM algorithm developed in this package is designed to augment. Slides are available on my [website](https://opmc2.github.io/project/ukheexp/). I will briefly summarise the content of the slides here, though I focus on the production of results and  corresponding plots in this vignette. I also use a different subsample to the slides.

# Research questions

- How important are earnings expectations for 16-18 year olds when deciding whether to go to university?
- How have these factors changed between the 1980s and today?
- What is driving the SES-gap in education attainment?

# Summary of what I do

- Model the (binary) decision to continue to university
- Estimate model separately for cohorts born in 1970 and 1990
  - Schooling, background, and subjective data from before decision
  - Earnings and occupation data after entry to labour market
  - Exploit subjective data on students’ attitudes and lifestyles at 16 to 
  better capture non-earnings factors
- Exploit model structure to give distributions of factors in decision
  - Compare distributions of earnings vs other factors within cohorts
  - Compare relative importance of factors across cohorts
  - Compare distributions for different demographics
- Use subjective data to decompose “psychic costs”
  - In progress
