---
title: "Probability model for role of non-pecuniary factors in decision to attend university"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{probModel}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
# library(ukheEm)
load_all()
library(ggplot2)

# load data
data("lsype1YP", "lsype1FB", "lsype1YPlabels", "lsype1FBlabels")

# create dtLsype
lsype1YP <- lapply(lsype1YP, as.data.table)
lsype1FB <- lapply(lsype1FB, as.data.table) 

benefitCols <- names(lsype1YP[[4]])[names(lsype1YP[[4]]) %like% "Benefits"]
costCols    <- names(lsype1YP[[4]])[names(lsype1YP[[4]]) %like% "Costs"]

WhyHeCols <- names(lsype1YP[[4]])[names(lsype1YP[[4]]) %like% "WHYHEYP"]
HeConCols <- names(lsype1YP[[4]])[names(lsype1YP[[4]]) %like% "HeCon"]

# new variable indicating if asked b/c questions
lsype1YP[[4]][, inSubSample := !is.na(W4BenefitsYP0a)]

```

```{r create dtLsype}

dtLsype <- merge(
  lsype1YP[[4]][, .(NSID, female = (W4SexYP == "Female"), ethGrp = W4ethgrpYP,
                    mainAct17 = W4MainActYP, inSubSample, 
                    numAlev17 = W4AlevNoYP,
                    plan2applyHe = fcase(
                      W4Heposs9YP %in% c("Very likely", "Fairly likely"), T,
                      W4Heposs9YP %in% c("Not very likely", 
                                         "Not at all likely"), F
                    ),
                    .SD), 
                .SDcols = c(benefitCols, costCols, WhyHeCols, HeConCols)], 
  lsype1YP[[8]][, .(NSID, W8USUAL, mainAct25 = W8DACTIVITY, 
                    employed25 = fcase(W8DWRK == "Currently employed", T,
                                       W8DWRK == "Not currently employed", F), 
                    degree = fcase(W8DDEGP == "First or higher degree", T,
                                   W8DDEGP == "No degree", F),
                    russellGrp = fcase(
                      W8DRUSSELL == "Awarded by Russell Group university", T,
                      W8DRUSSELL == "Awarded by other HE institution", F),
                    wage25 = W8GROW, everUni = (W8EVERUNI == "Yes"))],
  on = "NSID"
)

dtLsype <- merge(dtLsype, 
                 lsype1FB[[4]][, .(NSID, W4Inc1EstMP, 
                                   w4SOCMajorMP, w4SOCMajorSP,
                                   w4hiqualgMP, w4sexMP)])

setnames(dtLsype, 
         old = paste0(".SD.", c(benefitCols, costCols, WhyHeCols, HeConCols)),
         new = c(benefitCols, costCols, WhyHeCols, HeConCols))

```

# Next Steps (LSYPE1)

## New variables and labels

### Parental earnings: my measure of SES

```{r parInc}

dtLsype[, inc_midPts := fcase(
  W4Inc1EstMP == "Up to £2,599", 1300,
  W4Inc1EstMP == "£2,600 up to £5,199", 3900,
  W4Inc1EstMP == "£5,200 up to £10,399", 7500,
  W4Inc1EstMP == "£10,400 up to £15,599", 12500,
  W4Inc1EstMP == "£15,600 up to £20,799", 17500,
  W4Inc1EstMP == "£20,800 up to £25,999", 22500,
  W4Inc1EstMP == "£26,000 up to £31,199", 28000,
  W4Inc1EstMP == "£31,200 up to £36,399", 33000,
  W4Inc1EstMP == "£36,400 up to £41,599", 38000,
  W4Inc1EstMP == "£41,600 up to £46,799", 43000, 
  W4Inc1EstMP == "£46,800 up to £51,999", 48000,
  W4Inc1EstMP == "£52,000 or more", 55000
)]

dtLsype[, quantile(inc_midPts, probs = seq(0, 1, by = .05), na.rm = TRUE)]
dtLsype[inSubSample == T, quantile(inc_midPts, probs = seq(0, 1, by = .05), na.rm = TRUE)]

dtLsype[, parInc := fcase(
  W4Inc1EstMP %in% c("Up to £2,599", 
                     "£2,600 up to £5,199", 
                     "£5,200 up to £10,399", 
                     "£10,400 up to £15,599"), 1,
  W4Inc1EstMP %in% c("£15,600 up to £20,799",
                     "£20,800 up to £25,999",
                     "£26,000 up to £31,199",
                     "£31,200 up to £36,399",
                     "£36,400 up to £41,599",
                     "£41,600 up to £46,799"), 2,
  W4Inc1EstMP %in% c("£46,800 up to £51,999", 
                     "£52,000 or more"), 3
)]

dtLsype[, parInc := factor(parInc, levels = 1:3, 
                         labels = c("Bottom 20%",
                                    "Middle 60%",
                                    "Top 20%"))]

```

### Other factors ("psychic costs")

```{r other factor questions lsype}

# benefit labels
benefitLabels <- c(
  "Get better job", "Well-paid job", "Better opportunities", "Need for career",
  "Show skills", "Delay get job", "Social life", "Leave home", "Learning", 
  "More qualifications", "Personal development", "More confidence",
  "More respect", "Better life (general)", "Gain life skills", "Other", 
  "Don't know", "No answer"
)

costLabels <- c(
  "Expensive", "Get into debt", "Depend on parents", "Not financially indep.",
  "Not earning / working", "Costs (general)", "No job guarantee", 
  "Not needed for job", "Less experience", "Heavy workload", "Leave home",
  "Takes long time", "Waste of time", "Tuition fees etc.", "Stress", "Other",
  "Don't know", "No answer"
)

names(benefitLabels) <- benefitCols
names(costLabels) <- costCols

dtBenefits <- dtLsype[inSubSample == T, 
                            .(benefit = names(.SD), 
                              count = colSums(.SD == "Mentioned")), 
                            .SDcols = benefitCols, 
                      by = c("plan2applyHe", "degree")]

dtCosts <- dtLsype[inSubSample == T, 
                            .(cost = names(.SD), 
                              count = colSums(.SD == "Mentioned")), 
                            .SDcols = costCols, 
                   by = c("plan2applyHe", "degree")]

```

### Separating financial factors

```{r financial factors}

finLabels <- c(
  "Expensive", "Get into debt", "Depend on parents", "Not financially indep.",
  "Not earning / working", "Costs (general)", "Tuition fees etc."
)

finFactors <- names(costLabels)[costLabels %in% finLabels]

dtCosts[, highlight := fcase(
  cost %in% finFactors, 1,
  default = .2
)]

```


## Probability model of university attendance

The aim of this vignette is to carry out (and document) estimation of probability model(s) of university attendance including students answers to questions about the advantages and disadvantages of attending university. I initially focus on Next Steps data. I will use both realised attainment of a university degree (at age 25) and the intention to apply (at the time of the survey) as dependent variables.  

### Linear probability model (LPM)

The first specification is a (naive) linear probability model of the form
$$
D_i = X_i'\beta + Z_i'\gamma + \varepsilon_i
$$
where $D$ is the (binary) dependent variable (either attendance or intention to attend), $X$ are person characteristics (including ability, ethnicity, parental income), $Z$ are the advantages and disadvantages of attending university reported by the students, and $\epsilon$ is an error term.

```{r lpm attendance}

dtLsype[, .N, by = "mainAct17"]

xVars <- c(
  # YP characteristics
  "ethGrp", "female", "numAlev17",
  # Family background
  "parInc"
)

zVars <- c(benefitCols, costCols)

formula_degree_z <- as.formula(
  paste0('degree ~ ', 
         paste0(c(zVars), collapse = ' + '))
) 

formula_degree <- as.formula(
  paste0('degree ~ ', 
         paste0(c(xVars, zVars), collapse = ' + '))
)

lpm_degree_z <- lm(
  formula = formula_degree_z,
  data = dtLsype[mainAct17 == "Going to a school or college full time" &
                   inSubSample == T]
)

summary(lpm_degree_z)

dtLpmDegreeZ <- as.data.table(summary(lpm_degree_z)$coefficients, 
                             keep.rownames = TRUE)

dtLpmDegreeZ[, rn := stringr::str_remove(rn, "Mentioned")]

ggplot(data = dtLpmDegreeZ, 
       mapping = aes(
         x = rn, y = Estimate,
         ymin = Estimate - qt(.025, 2162, lower.tail = FALSE) * `Std. Error`,
         ymax = Estimate + qt(.025, 2162, lower.tail = FALSE) * `Std. Error`
       )) +
  geom_pointrange() +
  scale_x_discrete(name = "Variable", limits = c(benefitCols, costCols), 
                   labels = c(benefitLabels, costLabels)) +
  theme_bw() + theme(axis.text.x = element_text(angle=45, hjust  = 1), 
                     legend.position = c(.9, .15))

lpm_degree <- lm(
  formula = formula_degree,
  data = dtLsype[mainAct17 == "Going to a school or college full time" &
                   inSubSample == T]
)

summary(lpm_degree)

dtLpmDegree <- as.data.table(summary(lpm_degree)$coefficients, 
                             keep.rownames = TRUE)

dtLpmDegree[, rn := stringr::str_remove(rn, "Mentioned")]

ggplot(data = dtLpmDegree, 
       mapping = aes(
         x = rn, y = Estimate,
         ymin = Estimate - qt(.025, 2162, lower.tail = FALSE) * `Std. Error`,
         ymax = Estimate + qt(.025, 2162, lower.tail = FALSE) * `Std. Error`
       )) +
  geom_pointrange() +
  scale_x_discrete(name = "Variable", limits = c(benefitCols, costCols), 
                   labels = c(benefitLabels, costLabels)) +
  theme_bw() + theme(axis.text.x = element_text(angle=45, hjust  = 1), 
                     legend.position = c(.9, .15))

```

```{r lpm intention}

formula_plan2apply <- as.formula(
  paste0('degree ~ ', 
         paste0(c(xVars, zVars), collapse = ' + '))
)

lpm_plan2apply <- lm(
  formula = formula_plan2apply,
  data = dtLsype[mainAct17 == "Going to a school or college full time" &
                   inSubSample == T]
)

summary(lpm_plan2apply)

dtLpmPlan <- as.data.table(summary(lpm_plan2apply)$coefficients, 
                             keep.rownames = TRUE)

dtLpmPlan[, rn := stringr::str_remove(rn, "Mentioned")]

ggplot(data = dtLpmPlan, 
       mapping = aes(
         x = rn, y = Estimate,
         ymin = Estimate - qt(.025, 2162, lower.tail = FALSE) * `Std. Error`,
         ymax = Estimate + qt(.025, 2162, lower.tail = FALSE) * `Std. Error`
       )) +
  geom_pointrange() +
  scale_x_discrete(name = "Variable", limits = c(benefitCols, costCols), 
                   labels = c(benefitLabels, costLabels)) +
  theme_bw() + theme(axis.text.x = element_text(angle=45, hjust  = 1), 
                     legend.position = c(.9, .15))

```


### Logit model

Next I try a logit model which is generally a preferred specification for a binary dependent variable.

```{r logit attendance}

logit_degree <- glm(
  formula = formula_degree,
  family = binomial(),
  data = dtLsype[mainAct17 == "Going to a school or college full time" &
                   inSubSample == T]
)

summary(logit_degree)

dtLogitDegree <- as.data.table(summary(logit_degree)$coefficients, 
                             keep.rownames = TRUE)

dtLogitDegree[, rn := stringr::str_remove(rn, "Mentioned")]

ggplot(data = dtLogitDegree, 
       mapping = aes(
         x = rn, y = Estimate,
         ymin = Estimate - qt(.025, 2162, lower.tail = FALSE) * `Std. Error`,
         ymax = Estimate + qt(.025, 2162, lower.tail = FALSE) * `Std. Error`
       )) +
  geom_pointrange() +
  scale_x_discrete(name = "Variable", limits = c(benefitCols, costCols), 
                   labels = c(benefitLabels, costLabels)) +
  theme_bw() + theme(axis.text.x = element_text(angle=45, hjust  = 1), 
                     legend.position = c(.9, .15)) +
  coord_cartesian(ylim = c(-3,3))

# only z vars

logit_degreeZ <- glm(
  formula = formula_degree_z,
  family = binomial(),
  data = dtLsype[mainAct17 == "Going to a school or college full time" &
                   inSubSample == T]
)

summary(logit_degreeZ)

dtLogitDegreeZ <- as.data.table(summary(logit_degreeZ)$coefficients, 
                             keep.rownames = TRUE)

dtLogitDegreeZ[, rn := stringr::str_remove(rn, "Mentioned")]

ggplot(data = dtLogitDegreeZ, 
       mapping = aes(
         x = rn, y = Estimate,
         ymin = Estimate - qt(.025, 2162, lower.tail = FALSE) * `Std. Error`,
         ymax = Estimate + qt(.025, 2162, lower.tail = FALSE) * `Std. Error`
       )) +
  geom_pointrange() +
  scale_x_discrete(name = "Variable", limits = c(benefitCols, costCols), 
                   labels = c(benefitLabels, costLabels)) +
  theme_bw() + theme(axis.text.x = element_text(angle=45, hjust  = 1), 
                     legend.position = c(.9, .15)) +
  coord_cartesian(ylim = c(-3,3))

```
