---
title: "resultsLsypeEmParInc"
output: rmarkdown::html_vignette
bibliography: ukhe.bib
vignette: >
  %\VignetteIndexEntry{resultsBcsEm}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.dim = c(10,6),
  out.width = "100%",
  warning = FALSE
)
```

```{r setup}
# library(ukheEm)
library(ggplot2)
library(ggrepel)
library(here)
library(magrittr)

devtools::load_all()

# functions
signiff <- function(x, ...) if (is.numeric(x)) signif(x, digits = 3)
saveFig <- function(x, width = 10, height = 6) ggsave(here(figDir, x), width = width, height = height)

options(scipen = 15)

# directories
figDir <- "vignettes/fig/LsypeCog"
dir.create(here(figDir))

resLsype <- resLsypeCog

maxK <- length(resLsype)
```

This vignette analyses the results of running `progUkheEm()` on the Next Steps data. The algorithm is presented in detail in @cassagneau-francis_role_2021, and builds on the work in @cassagneau-francis_nonparametric_2020.

```{r combine results in big dt}

bigDt <- list()
for (K in 2:maxK) {
  bigDt[[K-1]] <- resLsype[[K]][[3]]
  bigDt[[K-1]][, nTypes := K]
}

bigDt <- rbindlist(bigDt)

# reorder types by mu
bigDt[, oldType := type]
# bigDt[, mu := Hmisc::wtd.mean(y1, pk), by = .(nTypes, type)]
bigDt[, type := frank(alpha1, ties.method = "dense"), by = .(nTypes)]
bigDt[, type := factor(type, levels = 1:18)]
setorder(bigDt, nTypes, type, id)

# save number of individuals
NN <- bigDt[nTypes == 2 & type == 1, .N]
```

# Choosing the number of types (K)

```{r dtLikeConverge}

dtLikeConverge <- list()
for (K in 2:maxK) {
  dtLikeConverge[[K-1]] <- data.table(
    nTypes = K,
    iter = 1:length(resLsype[[K]]$listLoglike),
    logLikelihood = unlist(resLsype[[K]]$listLoglike)
  )
}

dtLikeConverge <- rbindlist(dtLikeConverge)
dtLikeConverge[, maxIter := max(iter), by = nTypes]

```

## Convergence

```{r llhood convergence plots}

ggplot(data = dtLikeConverge,
       mapping = aes(x = iter, y = logLikelihood, colour = factor(nTypes))) +
  geom_line() +
  geom_label_repel(data = dtLikeConverge[iter == maxIter, ], 
            mapping = aes(x = iter, 
                          y = logLikelihood, label = paste0("K = ", nTypes)),
            nudge_x = 50, segment.alpha = .6) +
  # facet_wrap(vars(nTypes)) +
  scale_x_continuous(name = "Iteration") +
  scale_y_continuous(name = "Log-likelihood") +
  theme_classic() + theme(text = element_text(size = 15)) +
  guides(colour = "none")

saveFig("likelihoodConverge.png")
```

## Likelihood criteria

We use four criteria to help select the number of types:

- log-likelihood, $\ln\mathcal{L}$
- AIC ($\ln\mathcal{L} - j$) 
- BIC ($\ln\mathcal{L} - \frac{j}{2}\ln{N}$)
- ICL (see @biernacki_assessing_2000)

where $j$ is the number of estimated parameters and $N$ the number of observations. 

The parameters are:
$$
\Theta_K = \left\{\underset{2 \times K}{\mu(k), \sigma_{\nu}(k)}, \left\{ \underset{2 \times 2 \times K}{\alpha(k,d),\sigma_{\epsilon}(k,d)}, \left\{ \underset{4 \times K}{\pi(k,z,d)} \right\}_{z=0}^1\right\}_{d=0}^1 \right\}_{k=1}^K
$$
meaning $j = \#\Theta_K = (2 + 2 \times 2 + 4) \times K = 10K$

```{r likelihood criteria}

dtLikelihood <- dtLikeConverge[iter == maxIter, .(nTypes, iter, logLikelihood)]
dtLikelihood[, `:=` (
  AIC = logLikelihood - 10 * nTypes,
  BIC = logLikelihood - 10 * nTypes / 2 * log(NN)
)]

dtLikelihood <- melt(
  dtLikelihood,
  id.vars = c("nTypes", "iter"),
  variable.name = "criteria"
)

dtLikelihood[iter == 399, maxIter := TRUE]

ggplot(
  data = dtLikelihood,
  mapping = aes(x = nTypes, y = value, colour = criteria)
) +
  geom_line() +
  geom_point(aes(shape = maxIter), size = 1.5) +
  facet_wrap(vars(criteria), nrow = 2, scale = "free") +
  scale_x_continuous(name = "Number of types (G)", breaks = seq(2, maxK, by = 2)) +
  scale_y_continuous(name = NULL) +
  scale_shape_manual(
    name = NULL, labels = "Reached max. \niterations (400)",
    values = 1, breaks = TRUE
  ) +
  theme_classic() + 
  theme(legend.position = c(.75,.25), text = element_text(size = 15)) +
  guides(colour = "none")

saveFig("likelihoodCriteria.png")
```

## Posterior probabilities

```{r pk plot}

ggplot(
  data = bigDt[nTypes %between% c(2, 7), ], 
  mapping = aes(x = pk, fill = type)
) +
  geom_histogram() +
  facet_wrap(vars(nTypes), scale = "free") +
  scale_y_continuous(name = NULL) +
  theme_classic() +
  theme(text = element_text(size = 15)) +
  guides(fill = "none")

saveFig("pkPlotK2_K7.png")

```

## Size (and means) of groups

The groups sizes are reasonable, and the algorithm sorts strongly on ability. 

```{r groups size and means}

dtMu <- bigDt[, .(
  # P_k = sum(pk) / NN,
  alpha1 = Hmisc::wtd.mean(y1, pk),
  alpha1_mean = mean(mu),
  muCons = Hmisc::wtd.mean(w, pk)
), by = .(nTypes, type)]

dtPk <- bigDt[, .(P_k = sum(pk) / NN), by = .(nTypes, type, d)]

dtPk <- merge(dtPk, dtMu, by = c("nTypes", "type"))

dtPk[, alpha1Norm := (alpha1 - mean(alpha1)) / sd(alpha1)]
dtPk[, muConsNorm := (muCons - mean(muCons)) / sd(muCons)]

dtPk <- melt(
  dtPk, 
  measure.vars = c("alpha1Norm", "muConsNorm")
)

ggplot(
  data = dtPk[nTypes %between% c(2, 7)],
  mapping = aes(x = type)
) +
  geom_col(aes(y = P_k / 2, fill = d)) +
  geom_point(aes(y = (value + 4)/6, shape = variable)) +
  facet_wrap(vars(nTypes), 
             labeller = as_labeller(function(x) paste0("K = ", x))) +
  scale_x_discrete(name = "Type (k)") +
  scale_y_continuous(
    name = "Pr(k)",
    sec.axis = sec_axis(trans = ~. * 5 - 4)
  ) +
  scale_shape_discrete(name = NULL, 
                       limits = c("alpha1Norm", "muConsNorm"),
                       labels = c("Cog score", "Wage @25")) +
  scale_fill_discrete(name = NULL) +
  theme_classic() +
  theme(legend.position = "bottom", text = element_text(size = 15))

saveFig("groupSizesMeansK2_K7.png")

```

## Type-specific graduate-wage premium at 25

```{r wp} 

dtWagePrem <- bigDt[, .(
  alpha1 = Hmisc::wtd.mean(y1, pk),
  mu = Hmisc::wtd.mean(w, pk)
), by = .(nTypes, type, d)]

dtWagePrem <- dcast(dtWagePrem, 
                    formula = nTypes + type ~ d, 
                    value.var = c("alpha1", "mu"))

setnames(dtWagePrem, 
         old = c("alpha1_No degree", "alpha1_First or higher degree", 
                 "mu_No degree", "mu_First or higher degree"), 
         new = c("alpha_FALSE", "alpha_TRUE", "mu_FALSE", "mu_TRUE"))

dtWagePrem[, wagePrem := mu_TRUE - mu_FALSE]
dtWagePrem[, parIncPrem := alpha_TRUE - alpha_FALSE]

dtWagePrem <- merge(dtWagePrem, 
                    dtPk[variable == "alpha1Norm", .(nTypes, type, alpha1, P_k)],
                    by = c("nTypes", "type"))

# dtWagePrem <- melt(dtWagePrem, 
#                    measure.vars = c("testPrem", "wagePrem"))

ggplot(data = dtWagePrem[nTypes %between% c(2, 9)],
       mapping = aes(x = type, alpha = P_k)) +
  geom_col(aes(y = wagePrem, fill = "Wage", colour = "Wage")) +
  geom_col(aes(y = parIncPrem, fill = "Par. inc.", colour = "Par. inc.")) +
  facet_wrap(vars(nTypes), nrow = 2) +
  scale_x_discrete(name = "Type (k)") +
  scale_y_continuous(name = "% difference") +
  scale_fill_manual(name = "Premia",
                    values = c("red", NA),
                    limits = c("Wage", "Par. inc.")) +
  scale_colour_manual(name = "Premia",
                      values = c(NA, "blue"),
                      limits = c("Wage", "Par. inc.")) +
  scale_alpha_continuous(name = "Pr(k)") +
  theme_classic() +
  theme(text = element_text(size = 15))

saveFig("wagePremK2_K9.png")

ggplot(data = dtWagePrem[nTypes %between% c(8, 13)],
       mapping = aes(x = type, alpha = P_k)) +
  geom_col(aes(y = wagePrem, fill = "Wage", colour = "Wage")) +
  geom_col(aes(y = parIncPrem, fill = "Par. inc.", colour = "Par. inc.")) +
  facet_wrap(vars(nTypes)) +
  scale_x_discrete(name = "Type (k)") +
  scale_y_continuous(name = "% difference") +
  scale_fill_manual(name = "Premia",
                    values = c("red", NA),
                    limits = c("Wage", "Par. inc.")) +
  scale_colour_manual(name = "Premia",
                      values = c(NA, "blue"),
                      limits = c("Wage", "Par. inc.")) +
  scale_alpha_continuous(name = "Pr(k)") +
  theme_classic() +
  theme(text = element_text(size = 15))

saveFig("wagePremK8_K13.png")

# ggplot(data = dtWagePrem[nTypes %between% c(14, 18)],
#        mapping = aes(x = type, y = value, fill = variable, alpha = P_k)) +
#   geom_col(position = "dodge") +
#   facet_wrap(vars(nTypes)) +
#   theme_classic()

```
## Unconditional graduate wage premium

$$
WP = \sum_k \pi(k) WP(k), \,\,\, WP_{d=1} = \sum_k \pi(k|d=1) WP(k)
$$

```{r unconditional ATEs}

dtWPuncond <- merge(
  dtWagePrem,
  bigDt[d == "First or higher degree", .(
    P_kdT = sum(pk) / .N
  ), by = .(nTypes, type)],
  by = c("nTypes", "type")
) %>% 
  .[, .(
  WPu = sum(P_k * wagePrem),
  WPu_dT = sum(P_kdT * wagePrem)
), by = nTypes]


ggplot(data = dtWPuncond,
       mapping = aes(x = nTypes)) +
  geom_col(aes(y = WPu, fill = "ATE", colour = "ATE")) +
  geom_col(aes(y = WPu_dT, fill = "ATT", colour = "ATT")) +
  scale_x_continuous(name = "Number of types (K)") +
  scale_y_continuous(name = "", limits = c(0, .25)) +
  scale_fill_manual(name = "",
                    values = c("red", NA),
                    limits = c("ATE", "ATT")) +
  scale_colour_manual(name = "",
                      values = c(NA, "blue"),
                      limits = c("ATE", "ATT")) +
  theme_classic() +
  theme(text = element_text(size = 15), legend.position = c(.9, .9))

saveFig("uncondWP.png")

```

## Pr(k,d)

The bottom group remains a similar size as we increase K

```{r p_kd}

dtPkd <- bigDt[, .(
  P_kd = sum(pk) / NN,
  muD = Hmisc::wtd.mean(y1, pk, normwt = TRUE),
  sigmaNuD = sqrt(Hmisc::wtd.var(y1, pk, normwt = TRUE)),
  alpha = Hmisc::wtd.mean(y2, pk, normwt = TRUE),
  sigmaEps = sqrt(Hmisc::wtd.var(y2, pk, normwt = TRUE))
), by = .(nTypes, type, d)]

dtPkd[, muDNorm := (muD - mean(muD)) / sd(muD)]
dtPkd[, alphaNorm := (alpha - mean(alpha)) / sd(alpha)]

ggplot(
  data = dtPkd[nTypes %between% c(2, 5), ],
  mapping = aes(x = type, fill = d)
) +
  geom_col(aes(y = P_kd)) +
  facet_wrap(vars(nTypes), nrow = 2) +
  scale_y_continuous(name = "Pr(k,d)") +
  scale_fill_discrete(name = NULL) +
  theme_classic() +
  guides(shape = guide_legend()) +
  theme(text = element_text(size = 15), legend.position = "bottom")

saveFig("PrkdK2_K5.png", 7, 7)

ggplot(
  data = dtPkd[nTypes %between% c(2, 9), ],
  mapping = aes(x = type, fill = d)
) +
  geom_col(aes(y = P_kd)) +
  facet_wrap(vars(nTypes), nrow = 2) +
  scale_y_continuous(name = "Pr(k,d)") +
  scale_shape_manual(name = NULL,
                     breaks = c("Parental Inc. @16", "log-wage (@25)"),
                     values = c(21, 24),
                     labels = c("Parental Inc. @16", "log-wage (@25)")) +
  theme_classic() +
  guides(shape = guide_legend()) +
  theme(text = element_text(size = 15), legend.position = "bottom")

saveFig("PrkdK2_K9.png")

ggplot(
  data = dtPkd[nTypes %between% c(8, 13), ],
  mapping = aes(x = type, fill = d)
) +
  geom_col(aes(y = P_kd)) +
  facet_wrap(vars(nTypes), nrow = 2) +
  scale_y_continuous(name = "Pr(k,d)") +
  scale_shape_manual(name = NULL,
                     breaks = c("Parental Inc. @16", "log-wage (@25)"),
                     values = c(21, 24),
                     labels = c("Parental Inc. @16", "log-wage (@25)")) +
  theme_classic() +
  guides(shape = guide_legend()) +
  theme(text = element_text(size = 15))

saveFig("PrkdK8_K13.png")


```

## Means by d

```{r group mu by degree}

ggplot(
  data = dtPkd[nTypes %between% c(2, 9), ],
  mapping = aes(x = type, y = muD, fill = d, alpha = P_kd)
) +
  geom_col(position = "dodge") +
  # geom_point(aes(y = alpha , shape = "log-wage (@25)"), stroke = .3) +
  facet_wrap(vars(nTypes), nrow = 2) +
  scale_y_continuous(name = "Parental inc. (@16)") +
  scale_alpha_continuous(range = c(.3, 1)) +
  theme_classic() +
  guides(shape = guide_legend()) +
  theme(text = element_text(size = 15), legend.position = "bottom") +
  coord_cartesian(ylim = c(9, 12))

saveFig("MuByDK2_K9.png")

ggplot(
  data = dtPkd[nTypes %between% c(8, 13), ],
  mapping = aes(x = type, y = muD, fill = d)
) +
  geom_col(position = "dodge") +
  # geom_point(aes(y = alpha , shape = "log-wage (@25)"), stroke = .3) +
  facet_wrap(vars(nTypes), nrow = 2) +
  scale_y_continuous(name = "Parental inc. (@16)") +
  theme_classic() +
  guides(shape = guide_legend()) +
  theme(text = element_text(size = 15), legend.position = "bottom") +
  coord_cartesian(ylim = c(8, 12))

saveFig("MuByDK8_K13.png")

```

```{r group alpha by degree}

ggplot(
  data = dtPkd[nTypes %between% c(2, 5), ],
  mapping = aes(x = type, y = alpha, fill = d, alpha = P_kd)
) +
  geom_col(position = "dodge") +
  facet_wrap(vars(nTypes), nrow = 2) +
  scale_y_continuous(name = "Log-wage (@25)") +
  scale_alpha_continuous(name = "Pr(k,d)", range = c(.3, 1)) +
  scale_fill_discrete(name = NULL) + 
  theme_classic() +
  guides(shape = guide_legend()) +
  theme(text = element_text(size = 15), legend.position = "bottom") +
  coord_cartesian(ylim = c(5.8, 6.2))

saveFig("AlphaByDK2_K5.png", width = 7, height = 7)

ggplot(
  data = dtPkd[nTypes %between% c(2, 7), ],
  mapping = aes(x = type, y = alpha, fill = d, alpha = P_kd)
) +
  geom_col(position = "dodge") +
  facet_wrap(vars(nTypes), nrow = 2) +
  scale_y_continuous(name = "Log-wage (@25)") +
  scale_alpha_continuous(name = "Pr(k,d)", range = c(.3, 1)) +
  scale_fill_discrete(name = NULL) + 
  theme_classic() +
  guides(shape = guide_legend()) +
  theme(text = element_text(size = 15), legend.position = "bottom") +
  coord_cartesian(ylim = c(5.8, 6.2))

saveFig("AlphaByDK2_K7.png")

ggplot(
  data = dtPkd[nTypes %between% c(8, 13), ],
  mapping = aes(x = type, y = alpha, colour = d)
) +
  geom_point(shape = "triangle") +
  facet_wrap(vars(nTypes), nrow = 2) +
  scale_y_continuous(name = "Log-wage (@25)") +
  theme_classic() +
  guides(shape = guide_legend()) +
  theme(text = element_text(size = 15), legend.position = "bottom")

saveFig("AlphaByDK8_K13.png")

```

```{r group alpha by degree lines}

ggplot(
  data = dtPkd[nTypes %between% c(3, 5), ],
  mapping = aes(x = type, y = alpha, colour = d, 
                group = paste0(d, nTypes), linetype = factor(nTypes))
) +
  geom_line() +
  geom_text(data = dtPkd[nTypes == 5 & type == 5, ], 
                  aes(label = d), nudge_y = .02) +
  # facet_wrap(vars(nTypes), nrow = 2) +
  scale_y_continuous(name = "Log-wage (@25)") +
  theme_classic() +
  guides(colour = "none", linetype = "none") +
  theme(text = element_text(size = 15), legend.position = "bottom") +
  coord_cartesian(ylim = c(5.8, 6.25))

saveFig("AlphaByDK2_K9_lines.png")

ggplot(
  data = dtPkd[nTypes %between% c(8, 13), ],
  mapping = aes(x = type, y = alpha, colour = d)
) +
  geom_point(shape = "triangle") +
  facet_wrap(vars(nTypes), nrow = 2) +
  scale_y_continuous(name = "Log-wage (@25)") +
  theme_classic() +
  guides(shape = guide_legend()) +
  theme(text = element_text(size = 15), legend.position = "bottom")

saveFig("AlphaByDK8_K13.png")

```

## Analysing groups based on observables 

### Assigning types

```{r assign types}

bigDt[, maxPk := max(pk), by = .(nTypes, id)]

dtAssignedTypes <- bigDt[pk == maxPk, ]

```

### Assigned groups: means and sizes

```{r assigned group sizes and means}

dtATPkd <- dtAssignedTypes[, .(
  P_kd_AT = .N / NN,
  mu = mean(y1),
  alpha = mean(y2)
), by = .(nTypes, type, d)]

dtATPkd[, `:=` (
  muNorm = (mu - mean(mu)) / sd(mu),
  alphaNorm = (alpha - mean(alpha)) / sd(alpha)
)]

ggplot(
  data = dtATPkd[nTypes %between% c(2, 9), ],
  mapping = aes(x = type, fill = d)
) +
  geom_col(aes(y = P_kd_AT)) +
  facet_wrap(vars(nTypes), nrow = 2) +
  scale_y_continuous(name = "Pr(k,d)") +
  scale_shape_manual(name = NULL,
                     breaks = c("Parental Inc. @16", "log-wage (@25)"),
                     values = c(21, 24),
                     labels = c("Parental Inc. @16", "log-wage (@25)")) +
  theme_classic() +
  guides(shape = guide_legend()) +
  theme(text = element_text(size = 15), legend.position = "bottom")

saveFig("PrkdK2_K9_assgndTypes.png")

ggplot(
  data = dtATPkd[nTypes %between% c(2, 9), ],
  mapping = aes(x = type, y = mu, colour = d, shape = d)
) +
  geom_point() +
  facet_wrap(vars(nTypes), nrow = 2) +
  scale_y_continuous(name = "Log-wage (@25)") +
  theme_classic() +
  guides(shape = guide_legend()) +
  theme(text = element_text(size = 15), legend.position = "bottom")

saveFig("MuByDK2_K9_AT.png")

ggplot(
  data = dtATPkd[nTypes %between% c(2, 9), ],
  mapping = aes(x = type, y = alpha, colour = d, shape = d)
) +
  geom_point() +
  facet_wrap(vars(nTypes), nrow = 2) +
  scale_y_continuous(name = "Log-wage (@25)") +
  theme_classic() +
  guides(shape = guide_legend()) +
  theme(text = element_text(size = 15), legend.position = "bottom")

saveFig("AlphaByDK2_K9_AT.png")

```

### Merging with observed characteristics

```{r observed characteristics}

lsypeW8 <- as.data.table(lsype1YP[[8]])
dtLsypeWorkerFirm <- lsypeW8[, .(
  id = NSID, sex = W8CMSEX,
  # yrJobStart = b960270,
  inCouple = W8PART,
  married = W8CRNOWMA,
  cohabit = W8NRANY,
  kids = W8MORECH,
  nbKids = W8MCHMANY,
  health = W8GENA,
  SOC2010 = W8JOBDOSOCCODE,
  NSSEC5 = W8DNSSEC5,
  # soc, seg91, rgsc91, isco, goldth90, hgscale, # occupation / socio-econ classes
  # noVocQual = no_qual,
  # ol_ac, ol_de, gcse_ac, gcse_de, # O-level / GCSE grades
  # alevel, slevel, # A-level / S-level passes,
  # hed, pgce, opgc, postgrad, found, dip_un, other_ac, # academic qualifications
  # nvq_1, nvq_2, nvq_3, nvq_4, nvq_5, nvq_6, # nvqs at levels
  # ageLeftSchl = b960129, ageLeftFtEd = b960132,
  # trainCourse = b960134, yts = b960137, employerTrain = b960143, # traininig since end fte
  # lengthUnemp = b960258, 
  usualHoursWk = W8WRKHRS, 
  firmSize = W8EMP
)]

dtAssignedTypes <- merge(
  dtAssignedTypes[, .(nTypes, id, y1, y2, z, d, type, pk)], 
  dtLsypeWorkerFirm, 
  by = "id"
) %>% 
  merge(dtLsype4Em[, .(id = NSID, locScore, att2schlScr, ghqScr)])

```

```{r sumstats}

dtSumStats <- dtAssignedTypes[nTypes == 2, .(
  "Parental Inc. @16 (mean)" = mean(exp(y1)) / 52,
  "Par. Inc. (std dev.)" = sd(exp(y1) / 52),
  "Wage (mean)" = mean(exp(y2)),
  "Wage (std dev.)" = sd(exp(y2)),
  "Degree" = mean(d %like% "First or higher degree"),
  "Male" = mean(sex %like% "Male"),
  "In relationship" = mean(inCouple %like% "Yes"),
  "Married" = mean(married %in% c("Spouse")),
  # "Divorced" = mean(married %in% c("Separated", "Divorced")),
  "Live w/ partner" = mean(cohabit %like% "Yes"),
  "Has children" = mean(kids %like% "Yes"),
  "Num. of children" = mean(nbKids, na.rm = TRUE), 
  "In good health" = mean(health %in% c("Excellent", "Very good", "Good")),
  # "Occupation data" = mean(!is.na(hgscale)),
  # "H-G score" = mean(hgscale, na.rm = TRUE),
  # "Age left schl" = mean(as.numeric(paste0(ageLeftSchl)), na.rm = TRUE), 
  # "Age left ft ed." = mean(as.numeric(paste0(ageLeftFtEd)), na.rm = TRUE),
  # "Ever unemp." = mean(lengthUnemp %in% c(
  #   "3 months or less", "4-6 months", "7-11 months", "1-2 years", 
  #   "More than 2 years"
  # )),
  "Hours per week" = mean(usualHoursWk, na.rm = TRUE),
  "1--24" = mean(firmSize %like% "1-24"),
  "25--499" = mean(firmSize %like% "25-499"),
  "500 or more" = mean(firmSize %like% "500 or more"),
  "Attitude to schoo" = mean(att2schlScr, na.rm = TRUE),
  "Locus of control" = mean(locScore, na.rm = TRUE),
  "GHQ" = mean(ghqScr, na.rm = TRUE),
  .N
)] %>%
  t() %>%
  as.data.table(keep.rownames = TRUE)

xtable::xtable(dtSumStats, digits = 2)

# occupations
dtAssignedTypes[nTypes == 2, .(prop = .N / NN), by = .(NSSEC5)]

```


```{r table by type K2 to K4}

dtByTypeK2_4 <- dtAssignedTypes[nTypes %between% c(2, 4), .(
  "Parental Inc. @16 (mean)" = mean(exp(y1)) / 52,
  "Par. Inc. (std dev.)" = sd(exp(y1) / 52),
  "Wage (mean)" = mean(exp(y2)),
  "Wage (std dev.)" = sd(exp(y2)),
  "Degree" = mean(d %like% "First or higher degree"),
  "Male" = mean(sex %like% "Male"),
  "In relationship" = mean(inCouple %like% "Yes"),
  "Married" = mean(married %in% c("Spouse")),
  # "Divorced" = mean(married %in% c("Separated", "Divorced")),
  "Live w/ partner" = mean(cohabit %like% "Yes"),
  "Has children" = mean(kids %like% "Yes"),
  "Num. of children" = mean(nbKids, na.rm = TRUE), 
  "In good health" = mean(health %in% c("Excellent", "Very good", "Good")),
  # "Occupation data" = mean(!is.na(hgscale)),
  # "H-G score" = mean(hgscale, na.rm = TRUE),
  # "Age left schl" = mean(as.numeric(paste0(ageLeftSchl)), na.rm = TRUE), 
  # "Age left ft ed." = mean(as.numeric(paste0(ageLeftFtEd)), na.rm = TRUE),
  # "Ever unemp." = mean(lengthUnemp %in% c(
  #   "3 months or less", "4-6 months", "7-11 months", "1-2 years", 
  #   "More than 2 years"
  # )),
  "Hours per week" = mean(usualHoursWk, na.rm = TRUE),
  "1--24" = mean(firmSize %like% "1-24"),
  "25--499" = mean(firmSize %like% "25-499"),
  "500 or more" = mean(firmSize %like% "500 or more"),
  .N
), by = .(nTypes, type)]

setorder(dtByTypeK2_4, nTypes, type)

dtByTypeK2_4[, names(dtByTypeK2_4) := lapply(.SD, signiff), 
             .SDcols = names(dtByTypeK2_4)]

dtByTypeK2_4 <- as.data.table(t(dtByTypeK2_4), keep.rownames = TRUE)

xtable::xtable(dtByTypeK2_4, digits = 2)

knitr::kable(dtByTypeK2_4)
```

```{r table by type K5 to K6}

dtByTypeK5_6 <- dtAssignedTypes[nTypes %in% c(5, 6), .(
  "Parental Inc. @16 (mean)" = mean(exp(y1)) / 52,
  "Par. Inc. (std dev.)" = sd(exp(y1) / 52),
  "Wage (mean)" = mean(exp(y2)),
  "Wage (std dev.)" = sd(exp(y2)),
  "Degree" = mean(d %like% "First or higher degree"),
  "Male" = mean(sex %like% "Male"),
  "In relationship" = mean(inCouple %like% "Yes"),
  "Married" = mean(married %in% c("Spouse")),
  # "Divorced" = mean(married %in% c("Separated", "Divorced")),
  "Live w/ partner" = mean(cohabit %like% "Yes"),
  "Has children" = mean(kids %like% "Yes"),
  "Num. of children" = mean(nbKids, na.rm = TRUE), 
  "In good health" = mean(health %in% c("Excellent", "Very good", "Good")),
  # "Occupation data" = mean(!is.na(hgscale)),
  # "H-G score" = mean(hgscale, na.rm = TRUE),
  # "Age left schl" = mean(as.numeric(paste0(ageLeftSchl)), na.rm = TRUE), 
  # "Age left ft ed." = mean(as.numeric(paste0(ageLeftFtEd)), na.rm = TRUE),
  # "Ever unemp." = mean(lengthUnemp %in% c(
  #   "3 months or less", "4-6 months", "7-11 months", "1-2 years", 
  #   "More than 2 years"
  # )),
  "Hours per week" = mean(usualHoursWk, na.rm = TRUE),
  "1--24" = mean(firmSize %like% "1-24"),
  "25--499" = mean(firmSize %like% "25-499"),
  "500 or more" = mean(firmSize %like% "500 or more"),
  .N
), by = .(nTypes, type)]

setorder(dtByTypeK5_6, nTypes, type)

dtByTypeK5_6[, names(dtByTypeK5_6) := lapply(.SD, signiff), 
             .SDcols = names(dtByTypeK5_6)]

dtByTypeK5_6 <- as.data.table(t(dtByTypeK5_6), keep.rownames = TRUE)

xtable::xtable(dtByTypeK5_6, digits = 2)

knitr::kable(dtByTypeK5_6)
```


## Studying the "high" and "low" types

The following table includes only those individuals classified as the highest type for at least one value of $K$ between 3 and 5. 


```{r top types assigned types}

dtSumStats_topTypes <- dtAssignedTypes[type == nTypes & nTypes %between% c(3,5), .(
  "Parental Inc. @16 (mean)" = mean(exp(y1)) / 52,
  "Par. Inc. (std dev.)" = sd(exp(y1) / 52),
  "Wage (mean)" = mean(exp(y2)),
  "Wage (std dev.)" = sd(exp(y2)),
  "Degree" = mean(d %like% "First or higher degree"),
  "Male" = mean(sex %like% "Male"),
  "In relationship" = mean(inCouple %like% "Yes"),
  "Married" = mean(married %in% c("Spouse")),
  # "Divorced" = mean(married %in% c("Separated", "Divorced")),
  "Live w/ partner" = mean(cohabit %like% "Yes"),
  "Has children" = mean(kids %like% "Yes"),
  "Num. of children" = mean(nbKids, na.rm = TRUE), 
  "In good health" = mean(health %in% c("Excellent", "Very good", "Good")),
  # "Occupation data" = mean(!is.na(hgscale)),
  # "H-G score" = mean(hgscale, na.rm = TRUE),
  # "Age left schl" = mean(as.numeric(paste0(ageLeftSchl)), na.rm = TRUE), 
  # "Age left ft ed." = mean(as.numeric(paste0(ageLeftFtEd)), na.rm = TRUE),
  # "Ever unemp." = mean(lengthUnemp %in% c(
  #   "3 months or less", "4-6 months", "7-11 months", "1-2 years", 
  #   "More than 2 years"
  # )),
  "Hours per week" = mean(usualHoursWk, na.rm = TRUE),
  "1--24" = mean(firmSize %like% "1-24"),
  "25--499" = mean(firmSize %like% "25-499"),
  "500 or more" = mean(firmSize %like% "500 or more"),
  .N
), by = d] %>%
  .[, names(.) := lapply(.SD, signiff), .SDcols = names(.)] %>%
  t() %>%
  as.data.table(keep.rownames = TRUE)

xtable::xtable(dtSumStats_topTypes)

```

```{r low types assigned types}

dtAssignedTypes[, group := fcase(
  type == nTypes, factor(2, levels = 1:2, labels = c("Low type", "High type")),
  type == 1, factor(1, levels = 1:2, labels = c("Low type", "High type"))
)]

dtSumStats_hilowTypes <- dtAssignedTypes[!is.na(group) & nTypes %between% c(3,5), .(
  "Parental Inc. @16 (mean)" = mean(exp(y1)) / 52,
  "Par. Inc. (std dev.)" = sd(exp(y1) / 52),
  "Wage (mean)" = mean(exp(y2)),
  "Wage (std dev.)" = sd(exp(y2)),
  "Degree" = mean(d %like% "First or higher degree"),
  "Male" = mean(sex %like% "Male"),
  "In relationship" = mean(inCouple %like% "Yes"),
  "Married" = mean(married %in% c("Spouse")),
  # "Divorced" = mean(married %in% c("Separated", "Divorced")),
  "Live w/ partner" = mean(cohabit %like% "Yes"),
  "Has children" = mean(kids %like% "Yes"),
  "Num. of children" = mean(nbKids, na.rm = TRUE), 
  "In good health" = mean(health %in% c("Excellent", "Very good", "Good")),
  # "Occupation data" = mean(!is.na(hgscale)),
  # "H-G score" = mean(hgscale, na.rm = TRUE),
  # "Age left schl" = mean(as.numeric(paste0(ageLeftSchl)), na.rm = TRUE), 
  # "Age left ft ed." = mean(as.numeric(paste0(ageLeftFtEd)), na.rm = TRUE),
  # "Ever unemp." = mean(lengthUnemp %in% c(
  #   "3 months or less", "4-6 months", "7-11 months", "1-2 years", 
  #   "More than 2 years"
  # )),
  "Hours per week" = mean(usualHoursWk, na.rm = TRUE),
  "1--24" = mean(firmSize %like% "1-24"),
  "25--499" = mean(firmSize %like% "25-499"),
  "500 or more" = mean(firmSize %like% "500 or more"),
  "Attitude to school" = mean(att2schlScr, na.rm = TRUE),
  "Locus of control" = mean(locScore, na.rm = TRUE),
  "GHQ" = mean(ghqScr, na.rm = TRUE),
  .N
), by = .(group, d)] %>%
  .[order(group, d)] %>%
  # .[, names(.) := lapply(.SD, signif, digits = 4), .SDcols = names(.)] %>%
  t() %>%
  as.data.table(keep.rownames = TRUE)

xtable::xtable(dtSumStats_hilowTypes, digits = 2)

```

We can do the same exercise using the posterior probabilities directly, without assigning individuals to types.


```{r toptypes weighted pk}

dtSumStats_topTypesPk <- merge(
  bigDt[, .(id, nTypes, type, pk, y1, y2, d, z)], 
  dtLsypeWorkerFirm, 
  by = "id"
) %>%
  .[type == nTypes & nTypes %between% c(3,5), .(
  "Parental Inc. @16 (mean)" = Hmisc::wtd.mean(exp(y1), pk) / 52,
  "Par. Inc. (std dev.)" = sqrt(Hmisc::wtd.var(exp(y1) / 52, pk)),
  "Wage (mean)" = Hmisc::wtd.mean(exp(y2), pk),
  "Wage (std dev.)" = sqrt(Hmisc::wtd.var(exp(y2), pk)),
  "Degree" = Hmisc::wtd.mean(d %like% "First or higher degree", pk),
  "Male" = Hmisc::wtd.mean(sex %like% "Male", pk),
  "In relationship" = Hmisc::wtd.mean(inCouple %like% "Yes", pk),
  "Married" = Hmisc::wtd.mean(married %in% c("Spouse"), pk),
  # "Divorced" = mean(married %in% c("Separated", "Divorced")),
  "Live w/ partner" = Hmisc::wtd.mean(cohabit %like% "Yes", pk),
  "Has children" = Hmisc::wtd.mean(kids %like% "Yes", pk),
  "Num. of children" = Hmisc::wtd.mean(nbKids, pk, na.rm = TRUE), 
  "In good health" = Hmisc::wtd.mean(health %in% c("Excellent", "Very good", "Good"), pk),
  # "Occupation data" = mean(!is.na(hgscale)),
  # "H-G score" = mean(hgscale, na.rm = TRUE),
  # "Age left schl" = mean(as.numeric(paste0(ageLeftSchl)), na.rm = TRUE), 
  # "Age left ft ed." = mean(as.numeric(paste0(ageLeftFtEd)), na.rm = TRUE),
  # "Ever unemp." = mean(lengthUnemp %in% c(
  #   "3 months or less", "4-6 months", "7-11 months", "1-2 years", 
  #   "More than 2 years"
  # )),
  "Hours per week" = Hmisc::wtd.mean(usualHoursWk, pk, na.rm = TRUE),
  "1--24" = Hmisc::wtd.mean(firmSize %like% "1-24", pk),
  "25--499" = Hmisc::wtd.mean(firmSize %like% "25-499", pk),
  "500 or more" = Hmisc::wtd.mean(firmSize %like% "500 or more", pk),
  N = sum(pk)
), by = d] %>%
  .[, names(.) := lapply(.SD, signiff), .SDcols = names(.)] %>%
  t() %>%
  as.data.table(keep.rownames = TRUE)

xtable::xtable(dtSumStats_topTypesPk)

```


## Occupations of the high types

```{r occupations}

dtOccupationsPk <- merge(
  bigDt[, .(id, nTypes, type, pk, y1, y2, d, z)], 
  dtLsypeWorkerFirm, 
  by = "id"
) %>%
  .[, .(N_pk = sum(pk)), by = .(nTypes, type, d, NSSEC5)]

setorder(dtOccupationsPk, nTypes, type, d, NSSEC5)

dtOccupations <- dtAssignedTypes[, .N, by = .(nTypes, type, d, NSSEC5)]

setorder(dtOccupations, nTypes, type, d, NSSEC5)


dtOccupations <- merge(dtOccupations, dtOccupationsPk, 
                       by = c("nTypes", "type", "d", "NSSEC5"))

dtOccupations <- melt(dtOccupations, id.vars = c("nTypes", "type", "d", "NSSEC5"))

ggplot(data = dtOccupations[nTypes == type & nTypes %between% c(3, 5)],
       mapping = aes(x = paste0(d, " : ", variable), y = value, fill = NSSEC5)) +
  geom_col(position = "fill") +
  scale_x_discrete(labels = function(x) stringr::str_wrap(x, width = 8)) +
  theme_classic() +
  theme(text = element_text(size = 15))

saveFig("OccupationsTopTypes.png")
```


### Occupations of the high and low types

```{r occupations}

methodLabels <- c("Assigned types", "Weights")
names(methodLabels) <- c("N", "N_pk")
labelMethod <- function(x) methodLabels[x]

dtOccupations[, group := fcase(
  type == nTypes, factor(2, levels = 1:2, labels = c("Low type", "High type")),
  type == 1, factor(1, levels = 1:2, labels = c("Low type", "High type"))
)]

ggplot(data = dtOccupations[!is.na(group) & nTypes %between% c(3, 5)],
       mapping = aes(x = d, y = value, fill = NSSEC5)) +
  geom_col(position = "fill") +
  facet_grid(cols = vars(variable), rows = vars(group), 
             labeller = labeller(variable = as_labeller(labelMethod))) +
  scale_x_discrete(
    name = NULL,
    labels = function(x) stringr::str_wrap(x, width = 20)
  ) + 
  scale_y_continuous(name = NULL, labels = scales::percent) +
  scale_fill_discrete(name = NULL) +
  theme_classic() +
  theme(text = element_text(size = 15), legend.position = "bottom")

saveFig("OccupationsHiLoTypes.png")

dtAssignedTypes[, group := fcase(
  type == nTypes, factor(2, levels = 1:2, labels = c("Low type", "High type")),
  type == 1, factor(1, levels = 1:2, labels = c("Low type", "High type"))
)]
dtAssignedTypes[!is.na(NSSEC5) & !is.na(group) & nTypes %between% c(3, 5), grpSize := .N, by = .(group, d)]


as.data.table(t(dtAssignedTypes[!is.na(group) & nTypes %between% c(3, 5) & !is.na(NSSEC5), 
                .(prop = .N / unique(grpSize)), by = .(group, NSSEC5, d)][order(group, d, NSSEC5)]),
              keep.rownames = T)
```



