---
title: "resultsBcsEm"
output: rmarkdown::html_vignette
bibliography: ukhe.bib
vignette: >
  %\VignetteIndexEntry{resultsBcsEm}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.dim = c(10,6),
  out.width = "100%",
  warning = FALSE
)
```

```{r setup}
# library(ukheEm)
library(ggplot2)
library(ggrepel)
library(here)

devtools::load_all()

# functions
signif2 <- function(x, ...) if (is.numeric(x)) signif(x, digits = 2)
saveFig <- function(x, ...) ggsave(here(figDir, x), width = 10, height = 6)

options(scipen = 15)

resBcs <- resBcsTS

maxK <- length(resBcsTS)

# directories
figDir <- "vignettes/fig/BcsTS"
dir.create(here(figDir))
```

This vignette analyses the results of running `progUkheEm()` on the BCS data. The algorithm is presented in detail in @cassagneau-francis_role_2021, and builds on the work in @cassagneau-francis_nonparametric_2020.

```{r combine results in big dt}

bigDt <- list()
for (K in 2:maxK) {
  bigDt[[K-1]] <- resBcs[[K]][[3]]
  bigDt[[K-1]][, nTypes := K]
}

bigDt <- rbindlist(bigDt)

# reorder types by mu
bigDt[, oldType := type]
# bigDt[, mu := Hmisc::wtd.mean(y1, pk), by = .(nTypes, type)]
bigDt[, type := frank(mu, ties.method = "dense"), by = .(nTypes)]
bigDt[, type := factor(type, levels = 1:maxK)]
setorder(bigDt, nTypes, type, id)

# save number of individuals
NN <- bigDt[nTypes == 2 & type == 1, .N]
```

# Choosing the number of types (K)

```{r dtLikeConverge}

dtLikeConverge <- list()
for (K in 2:maxK) {
  dtLikeConverge[[K-1]] <- data.table(
    nTypes = K,
    iter = 1:length(resBcs[[K]]$listLoglike),
    logLikelihood = unlist(resBcs[[K]]$listLoglike)
  )
}

dtLikeConverge <- rbindlist(dtLikeConverge)
dtLikeConverge[, maxIter := max(iter), by = nTypes]

```

## Convergence

```{r llhood convergence plots}

ggplot(data = dtLikeConverge,
       mapping = aes(x = iter, y = logLikelihood, colour = factor(nTypes))) +
  geom_line() +
  geom_label_repel(data = dtLikeConverge[iter == maxIter, ], 
            mapping = aes(x = iter, 
                          y = logLikelihood, label = paste0("K = ", nTypes)),
            nudge_x = 50, segment.alpha = .6) +
  scale_x_continuous(name = "Iteration") +
  scale_y_continuous(name = "Log-likelihood") +
  theme_classic() + theme(text = element_text(size = 15)) +
  guides(colour = "none")

saveFig("likelihoodConverge.png")
```

## Likelihood criteria

We use four criteria to help select the number of types:

- log-likelihood, $\ln\mathcal{L}$
- AIC ($\ln\mathcal{L} - j$) 
- BIC ($\ln\mathcal{L} - \frac{j}{2}\ln{N}$)
- ICL (see @biernacki_assessing_2000)

where $j$ is the number of estimated parameters and $N$ the number of observations. 

The parameters are:
$$
\Theta_K = \left\{\underset{2 \times K}{\mu(k), \sigma_{\nu}(k)}, \left\{ \underset{2 \times 2 \times K}{\alpha(k,d),\sigma_{\epsilon}(k,d)}, \left\{ \underset{4 \times K}{\pi(k,z,d)} \right\}_{z=0}^1\right\}_{d=0}^1 \right\}_{k=1}^K
$$
meaning $j = \#\Theta_K = (2 + 2 \times 2 + 4) \times K = 10K$

```{r likelihood criteria}

dtLikelihood <- dtLikeConverge[iter == maxIter, .(nTypes, iter, logLikelihood)]
dtLikelihood[, `:=` (
  AIC = logLikelihood - 10 * nTypes,
  BIC = logLikelihood - 10 * nTypes / 2 * log(NN)
)]

dtLikelihood <- melt(
  dtLikelihood,
  id.vars = c("nTypes", "iter"),
  variable.name = "criteria"
)

dtLikelihood[iter == 399, maxIter := TRUE]

ggplot(
  data = dtLikelihood,
  mapping = aes(x = nTypes, y = value, colour = criteria)
) +
  geom_line() +
  geom_point(aes(shape = maxIter), size = 1.5) +
  facet_wrap(vars(criteria), nrow = 2, scale = "free") +
  scale_x_continuous(name = "Number of types (G)", breaks = seq(2, maxK, by = 2)) +
  scale_y_continuous(name = NULL) +
  scale_shape_manual(
    name = NULL, labels = "Reached max. \niterations (400)",
    values = 1, breaks = TRUE
  ) +
  theme_classic() + 
  theme(legend.position = c(.75,.25), text = element_text(size = 15)) +
  guides(colour = "none")

saveFig("likelihoodCriteria.png")
```

## Posterior probabilities

```{r pk plot}

ggplot(
  data = bigDt[nTypes %between% c(2, 7), ], 
  mapping = aes(x = pk, fill = type)
) +
  geom_histogram() +
  facet_wrap(vars(nTypes), scale = "free") +
  scale_y_continuous(name = NULL) +
  theme_classic() +
  theme(text = element_text(size = 15)) +
  guides(fill = "none")

saveFig("pkPlotK2_K7.png")

ggplot(
  data = bigDt[nTypes %between% c(8, 13), ], 
  mapping = aes(x = pk, fill = type)
) +
  geom_histogram() +
  facet_wrap(vars(nTypes), scale = "free") +
  scale_y_continuous(name = NULL) +
  theme_classic() +
  theme(text = element_text(size = 15)) +
  guides(fill = "none")

saveFig("pkPlotK8_K13.png")

ggplot(
  data = bigDt[nTypes %between% c(14, 18), ], 
  mapping = aes(x = pk, fill = type)
) +
  geom_histogram() +
  facet_wrap(vars(nTypes), scale = "free") +
  scale_y_continuous(name = NULL) +
  theme_classic() +
  theme(text = element_text(size = 15)) +
  guides(fill = "none")

saveFig("pkPlotK14_K18.png")

```

## Size (and means) of groups

The groups sizes are reasonable, and the algorithm sorts strongly on ability. 

```{r groups size and means}

dtPk <- bigDt[, .(
  P_k = sum(pk) / .N,
  mu = Hmisc::wtd.mean(y1, pk),
  mu_mean = mean(mu),
  alphaCons = Hmisc::wtd.mean(y2, pk)
), by = .(nTypes, type)]

dtPk[, muNorm := (mu - mean(mu)) / sd(mu)]
dtPk[, alphaConsNorm := (alphaCons - mean(alphaCons)) / sd(alphaCons)]

dtPk <- melt(
  dtPk, 
  measure.vars = c("muNorm", "alphaConsNorm")
)

ggplot(
  data = dtPk[nTypes %between% c(2, 7)],
  mapping = aes(x = type)
) +
  geom_col(aes(y = P_k / 2)) +
  geom_point(aes(y = (value + 4)/6, colour = variable)) +
  facet_wrap(vars(nTypes), 
             labeller = as_labeller(function(x) paste0("K = ", x))) +
  scale_x_discrete(name = "Type (k)") +
  scale_y_continuous(
    name = "Pr(k)",
    sec.axis = sec_axis(trans = ~. * 5 - 4)
  ) +
  scale_color_discrete(name = NULL, 
                       limits = c("muNorm", "alphaConsNorm"),
                       labels = c("Test score", "Wage @25")) +
  theme_classic() +
  theme(legend.position = "bottom", text = element_text(size = 15))

saveFig("groupSizesMeansK2_K7.png")


ggplot(
  data = dtPk[nTypes %between% c(8, 13)],
  mapping = aes(x = type)
) +
  geom_col(aes(y = P_k / 2)) +
  geom_point(aes(y = (value + 4)/6, colour = variable)) +
  facet_wrap(vars(nTypes), 
             labeller = as_labeller(function(x) paste0("K = ", x))) +
  scale_x_discrete(name = "Type (k)") +
  scale_y_continuous(
    name = "Pr(k)",
    sec.axis = sec_axis(trans = ~. * 5 - 4)
  ) +
  scale_color_discrete(name = NULL, 
                       limits = c("muNorm", "alphaConsNorm"),
                       labels = c("Test score", "Wage @25")) +
  theme_classic() +
  theme(legend.position = "bottom", text = element_text(size = 15))

saveFig("groupSizesMeansK8_K13.png")


ggplot(
  data = dtPk[nTypes %between% c(14, 18)],
  mapping = aes(x = type)
) +
  geom_col(aes(y = P_k / 2)) +
  geom_point(aes(y = (value + 4)/6, colour = variable)) +
  facet_wrap(vars(nTypes), 
             labeller = as_labeller(function(x) paste0("K = ", x))) +
  scale_x_discrete(name = "Type (k)") +
  scale_y_continuous(
    name = "Pr(k)",
    sec.axis = sec_axis(trans = ~. * 5 - 4)
  ) +
  scale_color_discrete(name = NULL, 
                       limits = c("muNorm", "alphaConsNorm"),
                       labels = c("Test score", "Wage @25")) +
  theme_classic() +
  theme(legend.position = "bottom", text = element_text(size = 15))

saveFig("groupSizesMeansK14_K18.png")

```

## Type-specific graduate wage-premium at 25

```{r wp} 

dtWagePrem <- bigDt[, .(
  alpha = Hmisc::wtd.mean(y2, pk),
  muD = Hmisc::wtd.mean(y1, pk)
), by = .(nTypes, type, d)]

dtWagePrem <- dcast(dtWagePrem, 
                    formula = nTypes + type ~ d, 
                    value.var = c("alpha", "muD"))

# setnames(dtWagePrem, old = c("FALSE", "TRUE"), new = c("alpha_dF", "alpha_dT"))

dtWagePrem[, wagePrem := alpha_TRUE - alpha_FALSE]
dtWagePrem[, testPrem := log(muD_TRUE) - log(muD_FALSE)]

dtWagePrem <- merge(dtWagePrem, 
                    dtPk[variable == "muNorm", .(nTypes, type, mu, P_k)],
                    by = c("nTypes", "type"))

ggplot(data = dtWagePrem[nTypes %between% c(2, 9)],
       mapping = aes(x = type, alpha = P_k)) +
  geom_col(aes(y = wagePrem, fill = "Wage", colour = "Wage")) +
  geom_col(aes(y = testPrem, fill = "Par. inc.", colour = "Test")) +
  facet_wrap(vars(nTypes), nrow = 2) +
  scale_x_discrete(name = "Type (k)") +
  scale_y_continuous(name = "% difference") +
  scale_fill_manual(name = "Premia",
                    values = c("red", NA),
                    limits = c("Wage", "Test")) +
  scale_colour_manual(name = "Premia",
                      values = c(NA, "blue"),
                      limits = c("Wage", "Test")) +
  scale_alpha_continuous(name = "Pr(k)") +
  theme_classic() +
  theme(text = element_text(size = 15))


saveFig("wagePremK2_K9.png")


ggplot(data = dtWagePrem[nTypes %between% c(8, 13)],
       mapping = aes(x = type, alpha = P_k)) +
  geom_col(aes(y = wagePrem, fill = "Wage", colour = "Wage")) +
  geom_col(aes(y = testPrem, fill = "Par. inc.", colour = "Test")) +
  facet_wrap(vars(nTypes), nrow = 2) +
  scale_x_discrete(name = "Type (k)") +
  scale_y_continuous(name = "% difference") +
  scale_fill_manual(name = "Premia",
                    values = c("red", NA),
                    limits = c("Wage", "Test")) +
  scale_colour_manual(name = "Premia",
                      values = c(NA, "blue"),
                      limits = c("Wage", "Test")) +
  scale_alpha_continuous(name = "Pr(k)") +
  theme_classic() +
  theme(text = element_text(size = 15))


saveFig("wagePremK8_K13.png")



```


## Pr(k,d)

The bottom group remains a similar size as we increase K

```{r p_kd}

dtPkd <- bigDt[, .(
  P_kd = sum(pk) / NN,
  muD = Hmisc::wtd.mean(y1, pk),
  sigmaNuD = sqrt(Hmisc::wtd.var(y1, pk, normwt = TRUE)),
  alpha = Hmisc::wtd.mean(y2, pk),
  sigmaEps = sqrt(Hmisc::wtd.var(y2, pk, normwt = TRUE))
), by = .(nTypes, type, d)]

dtPkd[, muDNorm := (muD - mean(muD)) / sd(muD)]
dtPkd[, alphaNorm := (alpha - mean(alpha)) / sd(alpha)]

ggplot(
  data = dtPkd[nTypes %between% c(2, 9), ],
  mapping = aes(x = type, fill = d)
) +
  geom_col(aes(y = P_kd)) +
  geom_point(aes(y = (muDNorm + 4)/6, shape = "Test score"), stroke = .3) +
  geom_point(aes(y = (alphaNorm + 4)/6, shape = "log-wage (@25)"), stroke = .3) +
  facet_wrap(vars(nTypes), nrow = 2) +
  scale_y_continuous(name = "Pr(k,d)") +
  scale_shape_manual(name = NULL,
                     breaks = c("Test score", "log-wage (@25)"),
                     values = c(21, 24),
                     labels = c("Test score", "log-wage (@25)")) +
  theme_classic() +
  guides(shape = guide_legend()) +
  theme(text = element_text(size = 15), legend.position = "bottom")

saveFig("PrkdK2_K9.png")

ggplot(
  data = dtPkd[nTypes %between% c(5, 10), ],
  mapping = aes(x = type, fill = d)
) +
  geom_col(aes(y = P_kd)) +
  geom_point(aes(y = (muDNorm + 4)/6, shape = "Test score"), stroke = .3) +
  geom_point(aes(y = (alphaNorm + 4)/6, shape = "log-wage (@25)"), stroke = .3) +
  facet_wrap(vars(nTypes), nrow = 2) +
  scale_y_continuous(name = "Pr(k,d)") +
  scale_shape_manual(name = NULL,
                     breaks = c("Test score", "log-wage (@25)"),
                     values = c(21, 24),
                     labels = c("Test score", "log-wage (@25)")) +
  theme_classic() +
  guides(shape = guide_legend())

ggplot(
  data = dtPkd[nTypes %between% c(11, 18), ],
  mapping = aes(x = type, fill = d)
) +
  geom_col(aes(y = P_kd)) +
  geom_point(aes(y = (muDNorm + 4)/6, shape = "Test score"), stroke = .3) +
  geom_point(aes(y = (alphaNorm + 4)/6, shape = "log-wage (@25)"), stroke = .3) +
  facet_wrap(vars(nTypes)) +
  scale_y_continuous(name = "Pr(k,d)") +
  scale_shape_manual(name = NULL,
                     breaks = c("Test score", "log-wage (@25)"),
                     values = c(21, 24),
                     labels = c("Test score", "log-wage (@25)")) +
  theme_classic() +
  guides(shape = guide_legend())

ggplot(
  data = dtPkd, 
  mapping = aes(x = type, colour = d)
) + 
  geom_point(aes(y = sigmaNuD)) +
  facet_wrap(vars(nTypes)) +
  theme_classic()
  
ggplot(
  data = dtPkd, 
  mapping = aes(x = type, colour = d)
) + 
  geom_point(aes(y = sigmaEps)) +
  facet_wrap(vars(nTypes)) +
  theme_classic()
```

## Analysing groups based on observables 

### Assigning types

```{r assign types}

bigDt[, maxPk := max(pk), by = .(nTypes, id)]

dtAssignedTypes <- bigDt[pk == maxPk, ]

```

### Assigned groups: means and sizes

```{r assigned group sizes and means}

dtATPkd <- dtAssignedTypes[, .(
  P_kd_AT = .N / NN,
  mu = mean(y1),
  alpha = mean(y2)
), by = .(nTypes, type, d)]

dtATPkd[, `:=` (
  muNorm = (mu - mean(mu)) / sd(mu),
  alphaNorm = (alpha - mean(alpha)) / sd(alpha)
)]

ggplot(
  data = dtATPkd[nTypes %between% c(2, 9), ],
  mapping = aes(x = type, fill = d)
) +
  geom_col(aes(y = P_kd_AT)) +
  geom_point(aes(y = (muNorm + 4)/6, shape = "Test score"), stroke = .3) +
  geom_point(aes(y = (alphaNorm + 4)/6, shape = "log-wage (@25)"), stroke = .3) +
  facet_wrap(vars(nTypes), nrow = 2) +
  scale_y_continuous(name = "Pr(k,d)") +
  scale_shape_manual(name = NULL,
                     breaks = c("Test score", "log-wage (@25)"),
                     values = c(21, 24),
                     labels = c("Test score", "log-wage (@25)")) +
  theme_classic() +
  guides(shape = guide_legend()) +
  theme(text = element_text(size = 15), legend.position = "bottom")

saveFig("PrkdK2_K9_assgndTypes.png")

```

### Merging with observed characteristics

```{r observed characteristics}

dtBcsWorkerFirm <- bcs70$bcs1996x[, .(
  id = bcsid, sex, degree,
  yrJobStart = b960270,
  inCouple = b960319,
  married = b960322,
  livingSitn = b960321,
  nbKids = b960333,
  age = b960338,
  health = b960432,
  soc, seg91, rgsc91, isco, goldth90, hgscale, # occupation / socio-econ classes
  noVocQual = no_qual,
  ol_ac, ol_de, gcse_ac, gcse_de, # O-level / GCSE grades
  alevel, slevel, # A-level / S-level passes,
  hed, pgce, opgc, postgrad, found, dip_un, other_ac, # academic qualifications
  nvq_1, nvq_2, nvq_3, nvq_4, nvq_5, nvq_6, # nvqs at levels
  ageLeftSchl = b960129, ageLeftFtEd = b960132,
  trainCourse = b960134, yts = b960137, employerTrain = b960143, # traininig since end fte
  lengthUnemp = b960258, 
  usualHoursWk = b960277, 
  firmSize = b960272
)]

dtAssignedTypes <- merge(
  dtAssignedTypes[, .(nTypes, id, y1, y2, z, d, type, pk)], 
  dtBcsWorkerFirm, 
  by = "id"
)

```


```{r table by type K2 to K4}

dtByTypeK2_4 <- dtAssignedTypes[nTypes %between% c(2, 4), .(
  "Test score (mean)" = mean(y1),
  "TS (std dev.)" = sd(y1),
  "Wage (mean)" = mean(exp(y2)),
  "Wage (std dev.)" = sd(exp(y2)),
  "Degree" = mean(d),
  "Male" = mean(sex %like% "Male"),
  "In relationship" = mean(inCouple %like% "Yes"),
  "Married" = mean(married %in% c("Married - 1st", "Remarried - 2nd+")),
  "Divorced" = mean(married %in% c("Separated", "Divorced")),
  "Live w/ partner" = mean(livingSitn %in% c("Living with spouse", 
                                             "Living as a couple")),
  "Has children" = mean(nbKids %in% paste0(1:5)),
  "Num. of children" = mean(as.numeric(paste0(nbKids)), na.rm = TRUE), 
  "In good health" = mean(health %in% c("Excellent", "Good")),
  "Occupation data" = mean(!is.na(hgscale)),
  "H-G score" = mean(hgscale, na.rm = TRUE),
  "Age left schl" = mean(as.numeric(paste0(ageLeftSchl)), na.rm = TRUE), 
  "Age left ft ed." = mean(as.numeric(paste0(ageLeftFtEd)), na.rm = TRUE),
  "Ever unemp." = mean(lengthUnemp %in% c(
    "3 months or less", "4-6 months", "7-11 months", "1-2 years", 
    "More than 2 years"
  )),
  "Hours per week" = mean(as.numeric(paste0(usualHoursWk)), na.rm = TRUE),
  "10 or less" = mean(firmSize %like% "10 or less"), 
  "11--25" = mean(firmSize %like% "11 - 25"),
  "26--99" = mean(firmSize %like% "26 - 99"),
  "100--499" = mean(firmSize %like% "100 - 499"),
  "500 or more" = mean(firmSize %like% "500 or more"),
  .N
), by = .(nTypes, type)]

setorder(dtByTypeK2_4, nTypes, type)

dtByTypeK2_4[, names(dtByTypeK2_4) := lapply(.SD, signif2), 
             .SDcols = names(dtByTypeK2_4)]

dtByTypeK2_4 <- as.data.table(t(dtByTypeK2_4), keep.rownames = TRUE)

xtable::xtable(dtByTypeK2_4, digits = 2)

knitr::kable(dtByTypeK2_4)
```

```{r table by type K5 to K6}

dtByTypeK5_6 <- dtAssignedTypes[nTypes %in% c(5, 6), .(
  "Test score (mean)" = mean(y1),
  "TS (std dev.)" = sd(y1),
  "Wage (mean)" = mean(exp(y2)),
  "Wage (std dev.)" = sd(exp(y2)),
  "Degree" = mean(d),
  "Male" = mean(sex %like% "Male"),
  "In relationship" = mean(inCouple %like% "Yes"),
  "Married" = mean(married %in% c("Married - 1st", "Remarried - 2nd+")),
  "Divorced" = mean(married %in% c("Separated", "Divorced")),
  "Live w/ partner" = mean(livingSitn %in% c("Living with spouse", 
                                             "Living as a couple")),
  "Has children" = mean(nbKids %in% paste0(1:5)),
  "Num. of children" = mean(as.numeric(paste0(nbKids)), na.rm = TRUE), 
  "In good health" = mean(health %in% c("Excellent", "Good")),
  "Occupation data" = mean(!is.na(hgscale)),
  "H-G score" = mean(hgscale, na.rm = TRUE),
  "Age left schl" = mean(as.numeric(paste0(ageLeftSchl)), na.rm = TRUE), 
  "Age left ft ed." = mean(as.numeric(paste0(ageLeftFtEd)), na.rm = TRUE),
  "Ever unemp." = mean(lengthUnemp %in% c(
    "3 months or less", "4-6 months", "7-11 months", "1-2 years", 
    "More than 2 years"
  )),
  "Hours per week" = mean(as.numeric(paste0(usualHoursWk)), na.rm = TRUE),
  "10 or less" = mean(firmSize %like% "10 or less"), 
  "11--25" = mean(firmSize %like% "11 - 25"),
  "26--99" = mean(firmSize %like% "26 - 99"),
  "100--499" = mean(firmSize %like% "100 - 499"),
  "500 or more" = mean(firmSize %like% "500 or more"),
  .N
), by = .(nTypes, type)]

setorder(dtByTypeK5_6, nTypes, type)

dtByTypeK5_6[, names(dtByTypeK5_6) := lapply(.SD, signif2), 
             .SDcols = names(dtByTypeK5_6)]

dtByTypeK5_6 <- as.data.table(t(dtByTypeK5_6), keep.rownames = TRUE)

xtable::xtable(dtByTypeK5_6, digits = 2)

knitr::kable(dtByTypeK5_6)
```














